---
title: "Iterators"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Learn about iteration.
- Iterators in base R.
- Iterators in purrr.
- Chapter 21 of [RDS](https://r4ds.had.co.nz/).
- [Purrr Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/purrr.pdf)

# For Loops

- Load the tidyverse

    ```{r, message=FALSE}
    library(tidyverse)
    ```


- *Iteration* is the repetition of some amount of code.

- If we didn't know the `sum()` function, how would we add up the elements
  of a vector?
  
    ```{r}
    x <- c(8, 1, 3, 1, 3)
    ```
  
- We could manually add the elements.
    ```{r}
    x[1] + x[2] + x[3] + x[4] + x[5]
    ```

    But this is prone to error (through copy and paste). Also, what if `x` has
    10,000 elements?

- For loops to the rescue!    
    ```{r}
    sumval <- 0
    for (i in seq_along(x)) {
       sumval <- sumval + x[[i]]
    }
    sumval
    ```
    
- Each for loop contains the following elements:

    1. **Output**: This is `sumval` above. We allocate the space for the output 
       *before* the for loop.
    2. **Sequence**: This is `seq_along(x)` above, which evaluates to `1 2 3 4 5`.
       These are the values that `i` will go through each iteration.
    3. **Body**: This is the code between the curly braces `{}`. This is the code
       that will be evaluated each iteration with a new value of `i`.

- In the above sequence, R internally transforms the code to:

    ```{r}
    sumval <- 0
    sumval <- sumval + x[[1]]
    sumval <- sumval + x[[2]]
    sumval <- sumval + x[[3]]
    sumval <- sumval + x[[4]]
    sumval <- sumval + x[[5]]
    sumval
    ```

- You often want to fill a vector with values. You should create this vector
  beforehand using the `vector()` function.

- For example, let's calculate a vector of cumulative sums of `x`.
    ```{r}
    cumvec <- vector(mode = "double", length = length(x))
    cumvec
    for (i in seq_along(cumvec)) {
       if (i == 1) {
         cumvec[[i]] <- x[[i]]
       } else {
         cumvec[[i]] <- cumvec[[i - 1]] + x[[i]] 
       }
    }
    cumvec
    
    ## Same as cumsum(x)
    cumsum(x)
    ```
  

- **Exercise**: The first two numbers of the 
  [Fibonacci Sequence](https://en.wikipedia.org/wiki/Fibonacci_number)
  are 0 and 1. Each succeeding number is the sum of the previous two numbers in 
  the sequence. For example, the third element is 1 = 0 + 1, while the fourth
  elements is 2 = 1 + 1, and the fifth element is 3 = 2 + 1. Use a for loop to
  calculate the first 100 Fibonacci Numbers. Sanity Check: The $\log_2$ of the 100th 
  Fibonacci Number is about 67.57.
  
    ```{r, echo = FALSE, eval = FALSE}
    fibvec <- vector(mode = "double", length = 100)
    for (i in seq_along(fibvec)) {
      if (i > 2) {
        fibvec[[i]] <- fibvec[[i - 1]] + fibvec[[i - 2]]
      } else if (i == 1) {
        fibvec[[i]] <- 0 
      } else if (i == 2) {
        fibvec[[i]] <- 1
      } else {
        stop(paste0("i = ", i))
      }
    }
    head(fibvec, n = 10)
    ```
  

- Looping is most often done over the columns of a data frame. 

- Note: for a 
  data frame `df`, `seq_along(df)` is the same as `1:ncol(df)` which is the
  same as `1:length(df)` (since data frames are special cases of lists).

- Let's calculate the mean of each column of `mtcars`
    ```{r}
    data("mtcars")
    mean_vec <- vector(mode = "numeric", length = length(mtcars))
    for (i in seq_along(mtcars)) {
      mean_vec[[i]] <- mean(mtcars[[i]], na.rm = TRUE)   
    }
    mean_vec
    colMeans(mtcars)
    ```
    
- Why not just use `colMeans()`? Well, there is no "`colSDs`" function, so iteration is 
  important for applying non-implemented functions to multiple elements in R.


- **Exercise**: Use a for loop to calculate the standard deviation of each plant
  trait in the `iris` data frame.
  
    ```{r, eval = FALSE, echo = FALSE}
    data("iris")
    sdvec <- rep(NA, length = 4)
    for (i in seq_along(sdvec)) {
      sdvec[i] <- sd(iris[[i]], na.rm = TRUE) 
    }
    sdvec
    ```
  

# purrr

## Basic Mappings

- R is a functional programming language. Which means that you can pass
  functions to functions.
  
- Suppose on `mtcars` we want to calculate the column-wise mean, the column-wise
  median, the column-wise standard deviation, the column-wise maximum, the 
  column-wise minimum, and the column-wise
  [MAD](https://en.wikipedia.org/wiki/Median_absolute_deviation). The for-loop
  would look very similar
  
    ```{r, eval=FALSE}
    funvec <- rep(NA, length = length(mtcars))
    for (i in seq_along(funvec)) {
      funvec[i] <- fun(mtcars[[i]], na.rm = TRUE) 
    }
    funvec
    ```
    
- Ideally, we would like to just tell R what function to apply to each column
  of `mtcars`. This is what the purrr package allows us to do.
  
- purrr is a part of the tidyverse, and so does not need to be loaded 
  separately.
  
- `map_*()` takes a vector (or list or data frame) as input, applies a provided
  function on each element of that vector, and outputs a vector of the same 
  length.
    
    - `map()` returns a list.
    - `map_lgl()` returns a logical vector.
    - `map_int()` returns an integer vector.
    - `map_dbl()` returns a double vector.
    - `map_chr()` returns a character vector.
  
    ```{r, eval = FALSE}
    map_dbl(mtcars, mean)
    map_dbl(mtcars, median)
    map_dbl(mtcars, sd)
    map_dbl(mtcars, mad)
    map_dbl(mtcars, min)
    map_dbl(mtcars, max)
    ```
  
- You can pass on more arguements in `map_*()`.

    ```{r, eval = FALSE}
    map_dbl(mtcars, mean, na.rm = TRUE)
    ```

- Suppose you want to get the output of `summary()` on each column.
    ```{r, eval = FALSE}
    map(mtcars, summary)
    ```

- **Exercise** (RDS 21.5.3.1): Write code that uses one of the map functions to:

    1. Determine the type of each column in `nycflights13::flights.`
    2. Compute the number of unique values in each column of `iris`.
    3. Generate 10 random normals for each of $\mu = -10, 0, 10, \ldots, 100$.
    
    ```{r, eval = FALSE, echo = FALSE}
    library(nycflights13)
    data("flights")
    data("iris")
    map_chr(flights, typeof)
    map_int(iris, function(x) length(unique(x)))
    map_dfc(seq(-10, 100, by = 10), rnorm, n = 10) ->
      temp
    ```

## Shortcuts

- You can refer to elements of the vector by "`.`" in a `map()` call if the `.f`
  argument is preceded by a "`~`". 
  For example, the following are three equivalent ways to calculate the mean
  of each column in `mtcars`.
  
    ```{r, eval = FALSE}
    map_dbl(mtcars, mean)
    map_dbl(mtcars, function(.) mean(.))
    map_dbl(mtcars, ~mean(.))
    ```
    
- What is actually going on is that purrr is creating an "anonymous function"

    ```{r, eval = FALSE}
    .f <- function(.) {
      mean(.)
    }
    ```

    and then calling this function in `map()`.

    ```{r, eval = FALSE}
    map_dbl(mtcars, .f)
    ```
  

- Why is this useful? Consider the following chunk of code which allows us to
  fit many simple linear regression models:

    ```{r}
    mtcars %>% 
      split(.$cyl) %>% 
      map(function(df) lm(mpg ~ wt, data = df)) ->
      lmlist
    ```
      
    - `split(.$cyl)` will turn the data frame into a list of data frames where
      each data frame has a different value of `cyl` for all units. The
      "`.`" references the current data frame.
    - `function(df) lm(mpg ~ wt, data = df)` defines a function (called an 
      "anonymous function") that will fit a linear model of `mpg` on `wt` 
      where those variables are in the data frame `df`.
    - The `map()` call fits that linear model to each of the three data frames
      in the list created by `split()`.
    - What is returned is a list of three `lm` objects that you can use to
      get fits and summaries.
      
    ```{r}
    summary(lmlist[[1]])
    ```
    
- Again, rather than create an "anonymous function", you can use the formula notation
  to do the same thing:
  
    ```{r}
    mtcars %>% 
      split(.$cyl) %>% 
      map(~lm(mpg ~ wt, data = .)) ->
      lmlist
    ```
  
    - Here, the "`.`" in "`data = .`" references the current data frame from
      the list of data frames that we are iterating through.
  
- We can use `map()` to get a list of summaries.
    ```{r}
    lmlist %>%
      map(summary) ->
      sumlist
    ```


- If you want to extract the $R^2$, you can do this using the formula notation
  as well.
  
    ```{r}
    sumlist[[1]]$r.squared ## only gets one R^2 out.
    
    ## Gets all R^2 out
    sumlist %>%
      map(~.$r.squared)
    ```
  
  
- **Exercise**: A [$t$-test](https://en.wikipedia.org/wiki/Student%27s_t-test)
  is used to test for differences in population means. R implements this
  with `t.test()`. For example, if I want to test for differences between
  the mean `mpg`'s of automatics and manuals (coded in variable `am`), I would
  use the follwoing syntax.
  
    ```{r, eval = FALSE}
    t.test(mpg ~ am, data = mtcars)$p.value
    ```
    
    Use `map()` to get the $p$-value for this test within each group of `cyl`.
    
    ```{r, eval = FALSE, echo = FALSE}
    mtcars %>%
      split(.$cyl) %>%
      map(~t.test(mpg ~ am, data = .)) %>%
      map(~.$p.value)
    ```

## `keep()` and `discard()`.

- `keep()` selects all variables that return `TRUE` according to some function.

- E.g. let's keep all numeric variables and calculate their means in the `iris`
  data frame.

    ```{r}
    iris %>%
      keep(is.numeric) %>%
      map_dbl(mean)
    ```
    
- `discard()` will select all variables that return `FALSE` according to some
  function.
  
- Let's count the number of each species.

    ```{r}
    iris %>%
      discard(is.numeric) %>%
      map(table)
    ```

- Other less useful functions are available in Section 21.9 of 
  [RDS](https://r4ds.had.co.nz/).


- **Exercise**: In the `mtcars` data frame, keep only variables that have a
  mean greater than `10` and calculate their mean. Hint: You'll have to use
  some of the shortcuts above.
  
    ```{r, eval = FALSE, echo = FALSE}
    mtcars %>%
      keep(~mean(.) > 10) %>%
      map(mean)
    ```



