---
title: "11 Vectors and Lists"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Manipulating Vectors and Lists using base R syntax.
- Chapter 20 of [RDS](https://r4ds.had.co.nz/).

# Vector Basics

- We'll use just a few tidyverse functions.
    ```{r, message=FALSE}
    library(tidyverse)
    ```

- Recall the vector material from the 
  [R Basics Worksheet](https://dcgerard.github.io/stat_412_612/lectures/01_intro/01_worksheet.pdf).

- A **vector** is a sequence of elements of the same type.

- "type" = `integer`, `double`, `character`, `logical`, `factor`, or `date`.

- `numeric` is used to represent vectors that are either integers or doubles.

- Use `c()` to create vectors.

- Use `typeof()` to see the type of vector and `is_*()` to check the type of 
  vector.

- Double:
    ```{r}
    x <- c(1, 10, 2)
    typeof(x)
    is_double(x) ## From purrr package
    ```

- Integer (use `L` to tell R that a number is an integer):

    ```{r}
    x <- c(1L, 10L, 2L)
    typeof(x)
    is_integer(x) ## From purrr package
    ```

- Character:
    ```{r}
    x <- c("hello", "good", "sir")
    typeof(x)
    is_character(x) ## From purrr package
    ```

- Logical:
    ```{r}
    x <- c(TRUE, FALSE, FALSE)
    typeof(x)
    is_logical(x) ## From purrr package
    ```

- Factor: Factors are actually integers with extra attributes.
    ```{r}
    x <- factor(c("A", "B", "B"))
    typeof(x)
    is.factor(x)
    is_logical(x) ## From purrr package
    ```
    
- Dates: Dates are actually doubles with extra attributes.
    ```{r}
    x <- lubridate::ymd(20150115, 20110630, 20130422)
    typeof(x)
    lubridate::is.Date(x)
    is_double(x) ## From purrr package
    ```

- Each element of a vector can have a name

    ```{r}
    x <- c(horse = 7, man = 1, dog = 8)
    x
    ```

- You can see and change the names with the `names()` function
    ```{r}
    names(x)
    names(x)[1] <- "cat"
    x
    ```

- Subset with brackets `[`
    ```{r}
    x <- c("I", "like", "dogs")
    x[2:3]
    lvec <- c(TRUE, FALSE, TRUE) 
    x[lvec]
    ```
    
- Substitute while subsetting

    ```{r}
    x[1] <- "You"
    x
    
    x[lvec] <- "We"
    x
    ```

- Subset with negative values to drop elements

    ```{r}
    x[-3]
    ```

- Subset a named vector with the name
    ```{r}
    x <- c(horse = 7, man = 1, dog = 8)
    x["man"]
    ```
    
- Two brackets `[[` only returns a single elements and drops the name.

    ```{r}
    x[3]
    x[[3]]
    ```
    
- **Exercise**: Consider the following vector:
    ```{r}
    x <- c(Yoshi = 10L,
           Mario = 31L,
           Luigi = 72L,
           Peach = 11L,
           Toad  = 38L)
    ```

    Extract Yoshi and Peach from the above vector using:

    1. Integer subsetting.
    2. Negative integer subsetting.
    3. Logical subsetting.
    4. Name subsetting.
    
    ```{r, echo = FALSE, eval = FALSE}
    x[c(1, 4)]
    x[c(-2, -3, -5)]
    x[c(TRUE, FALSE, FALSE, TRUE, FALSE)]
    x[c("Yoshi", "Peach")]
    ```

- **Exercise**: In the vector above, substitute Yoshi's number with `19L`.
    ```{r, eval = FALSE, echo = FALSE}
    x["Yoshi"] <- 19L
    x
    ```

    

- You are used to doing vectorized operations.
    ```{r}
    x <- c(1, 4, 1, 5)
    x + 10
    ```

- This is called "recycling", because what R is internally doing is thinking 
  this is the same as
    ```{r}
    x + c(10, 10, 10, 10)
    ```
    
- You can recycle non-scalars (but it's almost never a good idea):
    ```{r}
    x + c(10, 20)
    x + c(10, 20, 10, 20)
    ```


# Lists

- **Lists are vectors** whose elements can be of different types.

- Use `list()` to make a list.
    ```{r}
    my_first_list <- list(x = "a", y = 1, z = c(1L, 2L, 3L), list("a", 1))
    my_first_list                      
    ```

- The above is a *named* list that contains a character, a numeric, a logical vector, and another list. The internal list is *unnamed*.

- Use `str()` (for **str**ucture) to see the internal properties of a list.

    ```{r}
    str(my_first_list)
    ```


- Single brackets `[` return a sublist. You can use the same subsetting
  strategies as for vectors.

    ```{r}
    my_first_list[1:2]
    my_first_list["y"]
    ```
    
- Double brackets `[[` returns a single list element.
    ```{r}
    my_first_list[[1]]
    my_first_list[["z"]]
    ```
    
- Use dollar signs `$` (just like in data frames) to extract named list elements.

    ```{r}
    my_first_list$z
    ```
    
- You can remove elements of a list by substituting them with `NULL`.

    ```{r}
    str(my_first_list)
    my_first_list$x <- NULL
    str(my_first_list)
    ```



- **Exercise**: Consider the following list:

    ```{r}
    wedding <- list(venue = "chick-fil-a",
                    guest = tribble(~name,     ~meal, ~age,
                                    ##--------/------/-----
                                    "Yoshi",   "V",   29L,
                                    "Wario",   "C",   27L,
                                    "Bowser",  "V",   34L,
                                    "Luigi",   "C",   36L,
                                    "Toad",    "B",   34L), 
                    bride = "Peach",
                    groom = "Mario",
                    date  = parse_date("11/10/2020", "%d/%m/%Y"))
    ```
    
    a. Wario can't actually make it. Remove his row from the data frame.
    ```{r, eval = FALSE, echo = FALSE}
    wedding$guest %>%
      filter(name != "Wario") ->
      wedding$guest
    wedding$guest
    ```
    
    b. Add a new named vector called `meal` where `V` is `"Vegetarian"`, 
       `C` is `"Chicken"`, and `B` is `"Beef"`.
    ```{r, eval = FALSE, echo = FALSE}
    wedding$meal <- c(V = "Vegetarian", C = "Chicken", B = "Beef")
    wedding$meal
    ```
    
    c. Extract the venue and the date from `wedding`. Use three different
       techniques do this.
    ```{r, eval = FALSE, echo = FALSE}
    wedding[c(1, 5)]
    wedding[c("venue", "date")]
    wedding$venue
    wedding$date
    ```
    
    d. `"chick-fil-a"` should be capitalized. Capitalize the first `"c"`.
    
    ```{r, eval = FALSE, echo = FALSE}
    wedding$venue %>%
      str_replace(pattern = "^c", "C") ->
      wedding$venue
    wedding$venue
    ```
    
       
       
    

  







