---
title: "Vectors/Iterators Lab"
date: "`r Sys.Date()`"
author: "David Gerard"
output: pdf_document
urlcolor: "blue"
params: 
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$solutions, 
                      eval = params$solutions,
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Practice manipulating vectors.
- Practice for loops.
- Practice `purrr`.

# Exercise 1: Simulations

1. Write a function that uses a for loop that, for each iteration, randomly 
   draws 5 observations from an exponential distribution with rate parameter 1
   (use `rexp()`) and calculates its mean. It should do this 10,000 times. Choose
   an appropriate plot to plot the distribution of means.
   
    ```{r}
    maxit <- 10000
    n <- 5
    rate <- 1
    meanvec <- vector(mode = "numeric", length = maxit)
    for (i in seq_along(meanvec)) {
      meanvec[i] <- mean(rexp(n = n, rate = rate))
    }
    
    ggplot(data = tibble(meanvec), mapping = aes(x = meanvec)) +
      geom_histogram(bins = 30) +
      theme_bw() +
      xlab("Mean")
    ```
    
2. Repeat part 1 by using a `map_*()` function.

    ```{r}
    meanvec <- map_dbl(seq_len(maxit), ~mean(rexp(n = n, rate = rate)))
    
    ggplot(data = tibble(meanvec), mapping = aes(x = meanvec)) +
      geom_histogram(bins = 30) +
      theme_bw() +
      xlab("Mean")
    ```

3. Repeat part 1 by using the `replicate()` function.

    ```{r}
    meanvec <- replicate(n = maxit, expr = mean(rexp(n = n, rate = rate)))
    
    ggplot(data = tibble(meanvec), mapping = aes(x = meanvec)) +
      geom_histogram(bins = 30) +
      theme_bw() +
      xlab("Mean")
    ```

4. Use a another for loop that will print out plots for sample sizes of 5, 10,
   and 20 observations (instead of just 5).
   
    ```{r}
    nvec <- c(5, 10, 20)
    for (n in nvec) {
      meanvec <- replicate(n = maxit, expr = mean(rexp(n = n, rate = rate)))
      
      ggplot(data = tibble(meanvec), mapping = aes(x = meanvec)) +
        geom_histogram(bins = 30) +
        theme_bw() +
        xlab("Mean") +
        ggtitle(str_c("n = ", n)) ->
        pl
      print(pl)
    }
    ```
   

# Exercise 2: Star Trek

For this exercise, load in the following dataset.

```{r, message = FALSE, eval = TRUE, echo = TRUE}
library(tidyverse)
library(lubridate)
```

```{r, echo = TRUE, eval = TRUE}
st <- list(
  franchise = "Star Trek",
  series = c("TOS", "TNG", "DS9", "VOY", "ENT"),
  starship = tribble(~ID,         ~class,         ~name,       ~length,  ~launched,
                     ##----------/---------------/------------/---------/----------
                     "NCC-1701",  "constitution", "Enterprise", 288.646, 2245,
                     "NCC-1701D", "galaxy",       "Enterprise", 642.5,   2361,
                     "NX-74205",  "defiant",      "Defiant",    170,     2370,
                     "NCC-74656", "intrepid",     "Voyager",    343,     2371),
  captain = tribble(~firstname,  ~lastname, ~ship,       ~drink,
                    ##----------/----------/------------/----------
                    "James",     "Kirk",    "NCC-1701",  "brandy",
                    "Jean-luc",  "Picard",  "NCC-1701D", "tea",
                    "Benjamin",  "Sisko",   "NX-74205",  "coffee",
                    "Katherine", "Janeway", "NCC-74656", "coffee"),
  age = as.duration(today() - ymd(19660908))
)
```

1. Use an appropriate function to see the components of `st` in a concise way.

    ```{r}
    str(st)
    ```

2. Use R code to determine the class of each element of `st`.

    ```{r}
    map(st, class)
    ```
    
3. It seems that `starship` and `captain` can be merged into one data frame.
   Do this now.
   
    Your `st` list should now look like this:
    ```{r, eval = TRUE}
    st$starship %>%
      full_join(st$captain, by = c("ID" = "ship")) ->
      st$starship
    st$captain <- NULL
    str(st)
    ```
   
# Exercise 3: Diamonds

1. Load in the `diamonds` data frame from the ggplot2 package.

    ```{r, eval = TRUE}
    data("diamonds")
    ```

2. Calculate summary statistics (using `summary()`) for all quantitative
   variables.
   
    ```{r}
    diamonds %>%
      keep(is.numeric) %>%
      map(summary)
    ```
    
3. To fit a linear model of `price` on `depth` and extract the $p$-value for 
   the regression coefficient of `depth`, you would use the following R code:
   
    ```{r, echo = TRUE}
    lmout  <- lm(price ~ depth, data = diamonds)
    sumout <- summary(lmout)
    sumout$coefficients["depth", "Pr(>|t|)"]
    ```
    
    Fit a separate linear model for `price` on `depth` for each level of 
    `color`. Extract the $p$-values for the regression coefficients.
    
    ```{r}
    diamonds %>%
      split(.$color) %>%
      map(~lm(price ~ depth, data = .)) %>%
      map(~summary(.)) %>%
      map_dbl(~.$coefficients["depth", "Pr(>|t|)"]) %>%
      format(digits = 2, scientific = FALSE)
    ```

4. Plot `depth` vs `price`, faceting by `color`, including an OLS line. Do the
   $p$-values in the previous question roughly match what you would expect based
   on the plots?
   
    ```{r}
    ## No. It seems that a linear model isn't appropriate here. I bet we would see
    ## some weird residuals.
    library(hexbin)
    ggplot(diamonds, aes(x = depth, y = price)) +
      geom_hex() +
      facet_wrap(. ~ color) +
      geom_smooth(method = lm) +
      theme_bw()
    ```
   
   

