---
title: "Data Frames"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Based on a bootcamp originally written by Sohail Nizam.

# Learning Objectives

- Understand the fundamental type for storing data in R.
- Create and manipulate data frames (Tibbles), extract variables.
- Chapter 10 in [RDS](https://r4ds.had.co.nz/)

# Data Frames

- Usually we have more than one vector (variable) in our data set. 
- So how can we store several vectors together? 
- We'll use something called a data frame. 
- It's important to think of a data frame as a collection of columns,
  not a collection of rows because *that's how R thinks of it*. 
- When you have a data frame, you can easily refer to specific columns.
- But refering to rows becomes more complicated.
- Just like vectors are created with the `c()` function taking a collection of
  elements of the same type as input, data frames are created with the
  `data.frame()` function taking a collection of vectors as input. 
- The vectors can be of differing data types. 

- Let's play around with a dataframe from the `mtcars` dataset. To see a 
  description of these data, type
  
    ```{r, eval = FALSE}
    help(mtcars)
    ```

- We can load in these data with the `data` function

    ```{r}
    data("mtcars")
    ```



- If you have a relatively small data set, and you just want a cursory look at 
  the data, printing the data frame in the Console (my just typing `mtcars`)
  may suffice. However, if you have many columns and many rows, viewing your
  data in the console will be very difficult. 

- Instead, we can take a look at the data in a nice table in a new RStudio 
  tab using the `View()` function. 

- `View()` takes the name of a data frame as an argument. 
- Please note, `view()` is incorrect. The V must be capitalized.

    ```{r eval= FALSE}
    View(mtcars)
    ```


- More than likely, when you recieve data to work with, it will be in the form 
  of a data frame. 
- So once you have a data frame, how can you examine individual columns? 
- The syntax is very simple. To refer to one column simply type the name of 
  the data frame and the name of the column seperated by a $. For example:

    ```{r}
    mtcars$mpg #calls the mpg column of our data frame
    ```

- One nice thing about Rstudio is that it has a suggestion feature. 
- If you've saved a data frame, when you type its name and the dollar sign, a
  dropdown with all of the possible columns should appear for you. 
- If the dropdown does not appear, try pressing tab.

- Maybe you just want all of the column names displayed for you. 
- For that you can use the `names()` function. 
- `names()` takes a data frame as input and outputs a vector comprised of 
  that data frame's column names.

    ```{r}
    names(mtcars)
    ```

- Now let's use what we know about indexing to rename the first column. 
- We know that `names(mtcars)` represents a vector. 
- So let's refer to the first element of that vector and set it equal to
  something new. 

    ```{r}
    names(mtcars)[1] <- "mpg2" #rename the first column
    names(mtcars) #display the new names vector
    ```

- Here are some more useful functions for data frames:

    ```{r}
    head(mtcars, 15) #see the first 15 rows of the data frame
    tail(mtcars, 9) #see the last 9 rows of the data frams
    ```
    
# Tibbles

- The tidyverse uses tibbles more often than data frames.

- **Tibbles are mostly the same as data frames** with a few small exceptions:

    1. Better printing to console.
    2. Better interactions with strings.

- You can convert a data frame to a tibble with

    ```{r}
    suppressPackageStartupMessages(library(tidyverse))
    mtcars <- as_tibble(mtcars)
    mtcars
    ```
    
- **Exercise**: Extract the 8th to 28th elements of the `am` variable from the
  mtcars data frame.
  
    ```{r, echo = FALSE, eval = FALSE}
    mtcars$am[8:28]
    ```
  


