---
title: "Logicals"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives

- Boolean logic and R.
- Subsetting using logicals.

# Logicals

- A logical is a variable that can take on either `TRUE` or `FALSE`.

- Since 13 is greater than 4, the following is returned `TRUE`
    ```{r}
    13 > 4
    ```
    
    and the following is returned `FALSE`
    
    ```{r}
    13 < 4
    ```
    
- Use `>=` and `<=` to test for "greater than or equal" and "less than or equal", respectively

    ```{r}
    4 > 4
    4 >= 4
    4 < 4
    4 <= 4
    ```

- Use `==` comparisons to test if two quantities are equal:

    ```{r}
    13 == 4
    ```

- Use `!=` to test if two quantities are *not* equal:

    ```{r}
    13 != 4
    ```
    
- These operations can be vectorized:

    ```{r}
    x <- c(1, 2, 3, 4)
    y <- c(1, 4, 4, 4)
    x == y
    x != y
    x > y
    x < y
    ```
    
- Use "and" `&` to test if both of two conditions are `TRUE`

    ```{r}
    TRUE & TRUE
    TRUE & FALSE
    FALSE & TRUE
    FALSE & FALSE
    ```


- Use "or" `|` to test if either (or both) of two conditions are `TRUE`

    ```{r}
    TRUE | TRUE
    TRUE | FALSE
    FALSE | TRUE
    FALSE | FALSE
    ```
    
- `&` and `|` can also be vectorized:

    ```{r}
    x <- c(1, 2, 3, 4)
    y <- c(1, 4, 4, 4)
    (x < 3) & (y >= 4)
    (x < 3) | (y >= 4)
    ```
    
- Use logicals to extract elements of vectors

    ```{r}
    x <- 1:5
    x[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
    logvec <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
    x[logvec]
    logvec <- c(TRUE, FALSE, TRUE, TRUE, TRUE)
    x[logvec]
    ```


- Use logicals to extract elements of a vector that satisfy some condition
    ```{r}
    x <- 1:5
    logvec <- x < 3
    logvec
    x[logvec]
    ```
    
1. **Exercise**: If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Find the sum of all the multiples of 3 or 5 below 1000.

```{r, echo = FALSE, eval = FALSE}
int_vec <- 1:999
sum(int_vec[((int_vec %% 3) == 0) | ((int_vec %% 5) == 0)])
```

2. **Exercise**: What the sum of all integers that are either (divisible by 4 and less than 700) or (divisible by 3 and between 500 and 1000)?

```{r, echo = FALSE, eval = FALSE}
intvec <- 1:1000
(((intvec %% 4) == 0) & intvec < 700) &
  (((intvec %% 3) == 0) & intvec > 500) ->
  logvec
sum(intvec[logvec])
```



    

