---
title: "Logicals"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives

- Boolean logic and R.
- Subsetting using logicals.

# Logicals

- A logical is a variable that can take on either `TRUE` or `FALSE`.

- Since 13 is greater than 4, the following is returned `TRUE`
    ```{r}
    13 > 4
    ```
    
    and the following is returned `FALSE`
    
    ```{r}
    13 < 4
    ```
    
- Use `>=` and `<=` to test for "greater than or equal" and "less than or equal", respectively

    ```{r}
    4 > 4
    4 >= 4
    4 < 4
    4 <= 4
    ```

- Use `==` comparisons to test if two quantities are equal:

    ```{r}
    13 == 4
    ```

- Use `!=` to test if two quantities are *not* equal:

    ```{r}
    13 != 4
    ```
    
- These operations can be vectorized:

    ```{r}
    x <- c(1, 2, 3, 4)
    y <- c(1, 4, 4, 4)
    x == y
    x != y
    x > y
    x < y
    ```
    
- Use "and" `&` to test if both of two conditions are `TRUE`

    ```{r}
    TRUE & TRUE
    TRUE & FALSE
    FALSE & TRUE
    FALSE & FALSE
    ```


- Use "or" `|` to test if either (or both) of two conditions are `TRUE`

    ```{r}
    TRUE | TRUE
    TRUE | FALSE
    FALSE | TRUE
    FALSE | FALSE
    ```
    
- `&` and `|` can also be vectorized:

    ```{r}
    x <- c(1, 2, 3, 4)
    y <- c(1, 4, 4, 4)
    (x < 3) & (y >= 4)
    (x < 3) | (y >= 4)
    ```
    
\newpage

- Graphical Depiction of Logical Operations:

    ```{r, echo = FALSE, fig.height=4.5, fig.width=6}
    library(ggplot2)
    suppressPackageStartupMessages(library(gridExtra))
    plot_venn <- function(col1, col2, col3, col4, title, alpha = 0.5) {
        x1 <- seq(-1, 1, length = 300)
        y1 <- sqrt(1 - x1 ^ 2)
        y2 <- -sqrt(1 - x1 ^ 2)
        x2 <- x1 + 0.5
        x1 <- x1 - 0.5
        
        circdf1 <- data.frame(y = c(y1, y2[length(y2):1]), x = c(x1, x1[length(x1):1]))
        circdf2 <- data.frame(y = c(y1[length(y1):1], y2), x = c(x2[length(x2):1], x2))
        
        poly1 <- data.frame(x = c(x1[x1 < 0],
                                  x2[x2 < 0][sum(x2 < 0):1], 
                                  x2[x2 < 0], 
                                  x1[x1 < 0][sum(x1 < 0):1]),
                            y = c(y1[x1 < 0], 
                                  y1[length(y1):1][x2 < 0][sum(x2 < 0):1],
                                  y2[length(y2):1][x2 < 0],
                                  y2[x1 < 0][sum(x1 < 0):1]))
        poly2 <- poly1
        poly2$x <- poly1$x * -1
        
        poly3 <- data.frame(x = c(x2[x2 < 0][sum(x2 < 0):1], 
                                  x2[x2 < 0]),
                            y = c(y1[length(y1):1][x2 < 0][sum(x2 < 0):1],
                                  y2[length(y2):1][x2 < 0]))
        tempdf <- poly3[nrow(poly3):1, ]
        tempdf$x <- tempdf$x * -1
        poly3 <- rbind(poly3, tempdf)
        ggplot() +
          geom_polygon(data = poly1, 
                       mapping = aes(x = x, y = y),
                       fill = col1,
                       color = "black",
                       alpha = alpha) +
          geom_polygon(data = poly2, 
                       mapping = aes(x = x, y = y),
                       fill = col3,
                       color = "black",
                       alpha = alpha) +
          geom_polygon(data = poly3, 
                       mapping = aes(x = x, y = y),
                       fill = col2,
                       color = "black", 
                       alpha = alpha) +
          theme_void() +
          ggtitle(title) +
          theme(plot.title = element_text(hjust = 0.5), 
                plot.background = element_rect(color = col4)) +
          annotate(geom = "text", x = -0.8, y = 0, label = "x") +
          annotate(geom = "text", x = 0.8, y = 0, label = "y") ->
          pl
      return(pl)
    }
    
    aval <- 1
    col <- "#b1d9ef"
    grid.arrange(
      plot_venn(col, "white", "white", "white", "x & !y", alpha = aval),
      plot_venn("white", col, "white", "white", "x & y", alpha = aval),
      plot_venn("white", "white", col, "white", "!x & y", alpha = aval),
      plot_venn(col, col, "white", "white", "x", alpha = aval),
      plot_venn(col, "white", col, "white", "xor(x, y)", alpha = aval),
      plot_venn("white", col, col, "white", "y", alpha = aval),
      plot_venn(col, col, col, "white", "x | y", alpha = aval)
    )
    ```
    
- Use logicals to extract elements of vectors

    ```{r}
    x <- 1:5
    x[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
    logvec <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
    x[logvec]
    logvec <- c(TRUE, FALSE, TRUE, TRUE, TRUE)
    x[logvec]
    ```


- Use logicals to extract elements of a vector that satisfy some condition
    ```{r}
    x <- 1:5
    logvec <- x < 3
    logvec
    x[logvec]
    ```
    
1. **Exercise**: If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Find the sum of all the multiples of 3 or 5 below 1000.

```{r, echo = FALSE, eval = FALSE}
int_vec <- 1:999
sum(int_vec[((int_vec %% 3) == 0) | ((int_vec %% 5) == 0)])
```

2. **Exercise**: What the sum of all integers that are either (divisible by 4 and less than 700) or (divisible by 3 and between 500 and 1000)?

```{r, echo = FALSE, eval = FALSE}
intvec <- 1:1000
(((intvec %% 4) == 0) & intvec < 700) &
  (((intvec %% 3) == 0) & intvec > 500) ->
  logvec
sum(intvec[logvec])
```



    

