---
title: "Week 1 Lab"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

- Create a new R Markdown document. Number each of the following exercises. 
  Write your code solutions in code chunks. 
  
- Don't keep extraneous code. Only what is minimally sufficient to answer the 
  questions to stay in the file.

# Exercises

1. Sum up all even numbers from 10 to 1002 (look at the help page of `sum()` 
   and `seq()`)

```{r, echo = FALSE, eval = FALSE}
sum(seq(10, 1002, by = 2))
```

2. Create a vector of strings where the elements are `"Hello"`, `"world!"`, 
   `"How"`, `"are"`, and `"you?"`. Combine all of these words together into one
   string, separated by underscores (look at help page of `paste()`)
   
```{r, echo = FALSE, eval = FALSE}
string_vec <- c("Hello", "world!", "How", "are", "you?")
paste(string_vec, collapse = "_")
```
   
3. What happend when you sum `"1"` with `2`?

```{r, echo = FALSE, eval = FALSE, error=TRUE}
## You get an error
"1" + 2
```

4. You can comment out code using hashtags `#` at the start of a line. That is,
   the code will no longer run. Comments are great to make code more readable.
   Comment out the code in the previous chunk and rerun.
   
```{r, echo = FALSE, eval = FALSE}
## "1" + 2
```
   
5. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Find the sum of all the multiples of 3 or 5 below 1000.

This last one requires that you have some knowledge of logicals.

```{r, echo = FALSE, eval = FALSE}
int_vec <- 1:999
sum(int_vec[((int_vec %% 3) == 0) | ((int_vec %% 5) == 0)])
```


