---
title: "Common Coding Issues and How to Debug"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Practice reading common error messages.
- Learn good strategies to debugging.

# Strategies to Debug

- Run your code often so that you can spot errors early. Don't try to code everything at once and then run the whole thing. The earlier you spot a mistake, the easier it will be to debug.
- Read the error message very carefully. I know, there's a lot there and it's hard to weed out the important info from the excess text, but the more you practice the better you'll get at it!
- Run code one line at a time until you triangulate the location of the bug. Don't run the whole pipeline or ggplot sequence. Run it in parts.
- Do not run the whole chunk.
- Try restarting R with Session > Restart R. Sometimes, the error is caused because you changed something important (like redefining an important function by mistake).


- If you are having trouble Knitting:

    - Knit often so that you can spot errors early. The earlier you spot a bug the easier it will be to fix.
    - **Restart R** and *then* run the code one line at a time. When you knit, it will start from a blank R environment, so you want to reproduce that behavior to triangulate the bug.
    - Comment out sections of the document until it starts knitting. Uncomment sections until you find the issue. 

# Common Bugs

- Missing/extra Parentheses.
- Missing/extra commas
- Missing `+` (for ggplot)
- Missing pipe `%>%`
- Misspelled a function/variable name (using a lot of tab-completion helps limit the occurrence of this bug.)

- **Exercise**: Spot all of the errors below and fix the code.
    ```{r, eval = FALSE}
    library(tidyverse)
    data("mpg")
    ```

    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p", "r"))
      group_by(year) %>%
      summarize(mean_hwy = mean(hwy), 
                mean_cty = mean(cty))
      
    pl <- ggplot(mpg, aes(x = hwy, 
                          y = cty,
                          color = drv)) +
      geom_point() +
      geom_smooth(method = lm) +
      xlab("Highway MPG") +
      ylab(City MPG) +
      ggtitle("Highway vs City MPG")
    pl
    ```


    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p" "r")) %>%
      group_by(year) %>%
      summarize(mean_hwy = mean(hwy), 
                mean_cty = mean(cty))
      
    pl <- ggplot(mpg, aes(x = hwy, 
                          y = cty,
                          color = drv)) +
      geom_point() +
      geom_smooth(method = lm)
      xlab("Highway MPG") +
      ylab("City MPG") +
      ggtitle("Highway vs City MPG")
    pl
    ```
    
    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p", "r")) %>%
      group_by(year) %>%
      summarize(mean_hwy = mean(hwy), 
                mean_cty = mean(cty)
      
    pl <- ggplot(mpg, aes(x = hwy, 
                          y = cty,
                          color = drv)) +
      geom_point() +
      geom_smooth(method = lm) +
      xlab("Highway MPG") +
      ylab("City MPG") +
      ggtitle("Highway vs City MPG")
    pl
    ```
    
    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p", "r")) %>%
      group_by("year") %>%
      summarize(mean_hwy = mean(hwy), 
                mean_cty = mean(cty))
      
    pl <- ggplot(mpg, aes(x = hwy, 
                          y = cty
                          color = drv)) +
      geom_point() +
      geom_smooth(method = lm) +
      xlab("Highway MPG") +
      ylab("City MPG") +
      ggtitle("Highway vs City MPG")
    pl
    ```
    
    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p", "r")) %>%
      group_by(year) %>%
      summarize(mean_hwy = meanhwy), 
                mean_cty = mean(cty))
      
    pl <- ggplot(mpg, aes(x = hwy, 
                          y = cty,
                          color = drv)) +
      geom_point()) +
      geom_smooth(method = lm) +
      xlab("Highway MPG") +
      ylab("City MPG") +
      ggtitle("Highway vs City MPG")
    pl
    ```
    
    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p", "r")) %>%
      group_by(year) %>%
      summarize(mean_hwy  mean(hwy), 
                mean_cty = mean(cty))
      
    pl <- ggplot(mpg, aes(x = hwy, 
                          y = cty,
                          color = drv)) +
      geom_point() +
      geom_smooth(method = lm) ++
      xlab("Highway MPG") +
      ylab("City MPG") +
      ggtitle("Highway vs City MPG")
    pl
    ```
    
    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p", "r")) %>%
      group_by(year) %>%
      summarize(mean_hwy = mean(hwy), 
                mean_cty = mean(cty))
      
    pl <- ggplot(aes(x = hwy 
                     y = cty,
                     color = drv)) +
      geom_point() +
      geom_smooth(method = lm) +
      xlab("Highway MPG") +
      ylab("City MPG") +
      ggtitle("Highway vs City MPG")
    pl
    ```
    
    ```{r, error=TRUE, eval = FALSE}
    mpg %>%
      filter(fl %in% c("p", "r"))) %>%
      group_by(year) %>%
      summarize(mean_hwy = mean(hwy), 
                mean_cty = mean(cty))
      
    pl <- ggplot(mpg, aes(x = hwy, 
                          y = cty,
                          color = drv)) +
      geom_point() +
      geom_smooth(method = lm) +
      xlab("Highway MPG") +
      ylab("City MPG") +
      ggtitle("Highway vs City MPG""")
    pl
    ```
    
    




