---
title: "Data Import"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Import data from CSV's, 
- Working Directories
- Chapter 11 of [RDS](https://r4ds.had.co.nz/)

# Working Directories

- The working directory is where R will look for and save things by default.

- When you specify to save a figure, save a file, or load some data, it will be
  with respect to the working directory.
  
- You can see where the current working directory is by `getwd()`, or by looking
  at the top of the console in RStudio.

- You can change the working directory by Session > Set Working Directory > 
  Choose Directory. Or by CONTROL + SHIFT + HA. Or you can use the `setwd()` 
  command.
  
- A shortcut is to set the working directory to your sourcefile location with
  Session > Set Working Directory > To Source File Location.
  
- When you read and write files/figures, you can then specify the path
  from the position of the working directory.

- Suppose we want to save the following figure:  
    ```{r}
    suppressPackageStartupMessages(library(tidyverse))
    data("mpg")
    pl <- ggplot(mpg, aes(x = hwy, y = cty)) +
      geom_point()
    pl
    ```

- To save `pl` in the current folder, we would use:
    ```{r, eval = FALSE}
    ggsave(filename = "./my_saved_plot.pdf", plot = pl)
    ```
    
- The "`.`" means "the current folder".
    
- To save `pl` in the folder one level up we would use:
    ```{r, eval = FALSE}
    ggsave(filename = "../my_saved_plot.pdf", plot = pl)
    ```

- The "`..`" means "go one level up".

- If we are in the analysis folder, and we want to save `pl` in the output 
  folder, we would use:
  
    ```{r, eval = FALSE}
      ggsave(filename = "../output/my_saved_plot.pdf", plot = pl)
    ```
    
- If we have a subfolder called "fig" within out current folder. We could
  save `pl` in "fig" with
    ```{r, eval = FALSE}
      ggsave(filename = "./fig/my_saved_plot.pdf", plot = pl)
    ```
    
- **NEVER USE ABSOLUTE PATHS**. For example, you should never start the path
  from "C" if you use Windows. This makes your code non-transferable to other
  users.
  
# readr

- A lot of datasets come in comma-separated or tab-separated formats. For
  example, These are the first few rows of hate_crimes2.csv:
  
    ```{r, echo = FALSE, comment=""}
    writeLines(readLines(con = "../../data/hate_crimes2.csv", n = 10))
    ```

- In the file, each column is separated by a comma. Each row is separated by a
  new line.

- To read a CSV (comma-separated values) file into R, use the `read_csv()` 
  function from the readr package.
  
- The readr package is apart of the tidyverse, and so it is automatically
  loaded when you load the tidyverse.
  
    ```{r}
    library(tidyverse)
    hate_crimes <- read_csv(file = "../../data/hate_crimes1.csv")
    ```

- Use `read_tsv()` if columns are separted by tabs.

- Use `read_csv2()` if columns are separated by semi-colans.

- Other file formats are listed in [RDS](https://r4ds.had.co.nz/)

- You want to import data directly from Excel? Don't. 

    - First export the Excel spreadsheet as a CSV. Then read the CSV file into R.

- You are using colors to represent meaningful information in Excel? Don't.

    - Edit the data so that the information is encoded by a new variable.
    
# Special Considerations

- **Always check your data immediately after importing it**.

    - Check that the types are correct for each of the variables.
    - Check that the missing data were coded correctly.

- Sometimes the files code missing data other than `NA`. For example, it's common
  to use periods `.`, or in some genomic settings they use `-9` as missing.
  
- R won't know how to handle this without you telling it, so you'll have to 
  know what the missing data encoding is and specify it with the `na` argument
  in `read_csv()`.
  
- readr will try to guess the type for each column (double, integer, character,
  logic, etc). Sometimes it guesses wrong. If it seems to be guessing wong,
  use the `col_types` to explicitly specify the column types.
  
- Sometimes there are comments at the start of a data file. You can skip the 
  first few lines before starting to read data with the `skip` argument.
  
- If the comments begin with a special character, you can use the `comment` 
  argument.
  
# 

  

