---
title: "Housing EDA"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
params:
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = params$solutions, 
                      eval       = params$solutions,
                      fig.align  = "center",
                      fig.height = 3, 
                      fig.width  = 5)
```

Researchers were interested in predicting residential home sales prices in a 
Midwestern city as a function of various characteristics of the home and 
surrounding property. Data on 522 transactions were obtained for home sales 
during the year 2002. The 13 variables are

- `Price`: Sales price of residence (in dollars)
- `Area`: Finished area of residence (in square feet)
- `Bed`: Total number of bedrooms in residence
- `Bath`: Total number of bathrooms in residence
- `AC`: 1 = presence of air conditioning, 0 = absence of air conditioning
- `Garage`: Number of cars that a garage will hold
- `Pool`: 1 = presence of a pool, 0 = absence of a pool
- `Year`: Year property was originally constructed
- `Quality`: Index for quality of construction. `High`, `Medium`, or `Low`.
- `Style`: Categorical variable indicating architectural style 
- `Lot`: Lot size (in square feet)
- `Highway`: 1 = highway adjacent, 0 = highway not adjacent.

The data are available in "estate.csv" at https://dcgerard.github.io/stat_412_612/data/estate.csv.

Peform an exploratory data analysis to come up with some hypotheses. Some 
suggested ways to focus your research:

- Wich variables are categorical? Which are quantitative?
- What variables are marginally associated with price? Use plots and summary
  statistics.
- What variables are marginally associated with each other? Use plots and 
  summary statistics.
- If a variable is marginally associated with price, are there some other
  variables that could explain that association? Use plots and 
  summary statistics.
- Does there appear to be any discrete groupings of houses?
- Are there any unusual observations?
- What transformations should you perform to make associations more linear?
- What variables should be discretized (or have values aggregated) because
  there are too few values and/or the association seems discrete?


```{block}
Load required packages
```
```{r}
library(tidyverse)
library(GGally) ## for pairs plot
```

```{block}
Load Data into R
```
```{r}
estate <- read_csv("../../data/estate.csv")
```


```{r, message=FALSE}
estate %>%
  select(Price, Area, Year, Lot) %>%
  ggpairs()
```


