---
title: "Exploratory Data Analysis (EDA) in R"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Strategies for EDA
- Chapter 7 of [RDS](https://r4ds.had.co.nz/)

# General Strategies

- Plot the distribution of every variable.
- Plot the bivariate distribution of every pair of variables (to find which
  variables are associated).
- Color code by variables to try and see if relationships can be explained.
- Calculate lot sof summary statistics.
- Look at missingness.
- Look at outliers.
- EDA is about **curiosity**. Ask *many* questions, use *many* plots, 
  investigate *many* aspects of your data. This will let you hone in on 
  the few *interesting* questions you want to pursue deeper.
  
```{r, message=FALSE}
library(tidyverse)
data("diamonds")
```

  
# Distribution of Every Variable:

- Quantitative: Use a histogram.
    
    - Look for modality. Indicates multiple groups of units. What can explain 
      the modes? Can any of the other variables explain the modes?
    - Are certain values more likely than other values?
    - Look for skew. 
    - Does it make sense to think of these data on the log scale? For example, 
      if you look at two adjacent bins of (i) [1,2) and [2,3) or (ii) [1, 10) 
      and [10, 100). You should log in case (ii) and not log in case (i).
    - `geom_histogram()`
    - [Mean](https://en.wikipedia.org/wiki/Mean), 
      [median](https://en.wikipedia.org/wiki/Median), 
      [standard deviation](https://en.wikipedia.org/wiki/Standard_deviation),
      [five number summary](https://en.wikipedia.org/wiki/Five-number_summary).

    ```{r}
    ggplot(data = diamonds, mapping = aes(x = carat)) +
      geom_histogram(bins = 500)
    
    fivenum(diamonds$carat)
    mean(diamonds$carat)
    sd(diamonds$carat)
    ```

- Categorical: Use a bar chart. Or just a table of *proportions* (`table()` then
  `prop.table()`).

    - Absolute counts are sometimes interesting, but usually you want to 
      look at the proportion of observations in each category.
    - Is there a natural ordering of the categories (bad, medium, good)?
    - Why are some categories more represented than others?
    - `geom_bar()`, `geom_col()`
    - Proportion of observations within each group.
    
    ```{r}
    ggplot(diamonds, aes(x = color, y = ..)) +
      geom_bar(aes(y = ..count.. / sum(..count..))) +
      ylab("Proportion")
    
    table(diamonds$color)
    prop.table(table(diamonds$color))
    ```

# Bivariate Distribution of Every Pair of Variables

- Quantitative vs Quantitative: Use a scatterplot
    
    - Is the relationship linear? Quadratic? Exponential?
    - Logging is useful tool to make some associations linear. 
      If the relationship is (i) monotonic and (ii) curved, then try logging the x-variable.
      If it is also (iii) more variable at larger y-values, then try logging
      the y-variable *instead* of the x-variable. Try logging both if you still
      see curvature.
    - Ask if an observed association can be explained by another variable?
    - Correlation coefficient. Kendall's tau.
    
    ```{r}
    ggplot(diamonds, aes(x = carat, y = price)) +
      geom_point() +
      scale_y_log10() +
      scale_x_log10()
    
    cor(diamonds$carat, diamonds$price)
    ## cor(diamonds$carat, diamonds$price, method = "kendall")
    ```
    
- Categorical vs Quantitative: Use a boxplot

    - For which levels of the categorical variable is the quantitative variable
      higher or lower?
    - For which levels is the quantitative variable more spread out?
    - Aggregated means, medians, standard deviations, quantiles
    
    ```{r}
    ggplot(diamonds, aes(x = color, y = price)) +
      geom_boxplot() +
      scale_y_log10()
    
    diamonds %>%
      mutate(logprice = log(price)) %>%
      group_by(color) %>%
      summarize(mean   = mean(logprice), 
                sd     = sd(logprice), 
                median = median(logprice),
                Q1     = quantile(logprice, 0.25), 
                Q3     = quantile(logprice, 0.75))
    ```
    

- Categorical vs Categorical: Use a mosaic plot or a count plot

    - For which pairs of categorical variables are there the most number of 
      units?
    - Does the conditional distribution of a categorical variable change
      at different levels of the other categorical variable?
    - `prop.table()`

    ```{r}
    ## Only gives you the bivariate distribution
    ggplot(diamonds, aes(x = cut, y = color)) +
      geom_count()
    
    ## Gives you the conditional distributions of color given cut
    ggplot(diamonds, aes(x = cut, fill = color)) +
      geom_bar(position = "fill")
    
    ## Gives you the conditional distributions of cut given color
    ggplot(diamonds, aes(x = color, fill = cut)) +
      geom_bar(position = "fill")
    
    ## Bivariate Distribution
    prop.table(table(diamonds$color, diamonds$cut))
    
    ##  Conditional distributions of column variable conditional on row variable
    prop.table(table(diamonds$color, diamonds$cut), margin = 1)
    
    ## Conditional distributions of row variable conditional on column variable
    prop.table(table(diamonds$color, diamonds$cut), margin = 2)
    ```














