---
title: "Housing EDA"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
params:
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = params$solutions, 
                      eval       = params$solutions,
                      fig.align  = "center",
                      fig.height = 3, 
                      fig.width  = 5)
```

Researchers were interested in predicting residential home sales prices in a 
Midwestern city as a function of various characteristics of the home and 
surrounding property. Data on 522 transactions were obtained for home sales 
during the year 2002. The 13 variables are

- `Price`: Sales price of residence (in dollars)
- `Area`: Finished area of residence (in square feet)
- `Bed`: Total number of bedrooms in residence
- `Bath`: Total number of bathrooms in residence
- `AC`: 1 = presence of air conditioning, 0 = absence of air conditioning
- `Garage`: Number of cars that a garage will hold
- `Pool`: 1 = presence of a pool, 0 = absence of a pool
- `Year`: Year property was originally constructed
- `Quality`: Index for quality of construction. `High`, `Medium`, or `Low`.
- `Style`: Categorical variable indicating architectural style 
- `Lot`: Lot size (in square feet)
- `Highway`: 1 = highway adjacent, 0 = highway not adjacent.

The data are available in "estate.csv" at https://dcgerard.github.io/stat_412_612/data/estate.csv.

Perform an exploratory data analysis to come up with some hypotheses. Some 
suggested ways to focus your research:

- Which variables are categorical? Which are quantitative?
- Change the values of the categorical variables to something more informative.
- What variables are marginally associated with price? Use plots and summary
  statistics.
- What variables are marginally associated with each other? Use plots and 
  summary statistics.
- If a variable is marginally associated with price, are there some other
  variables that could explain that association? Use plots and 
  summary statistics.
- Does there appear to be any discrete groupings of houses?
- Are there any unusual observations?
- What transformations should you perform to make associations more linear?
- Try making new variables based on existing variables.
- What variables should be discretized (or have values aggregated) because
  there are too few values and/or the association seems discrete?
- If you know linear regression, try out the `lm()` and `step()` functions to 
  choose a tentative model.


```{block}
Load required packages
```
```{r}
library(tidyverse)
library(GGally) ## for pairs plot
theme_set(theme_bw())
```

```{block}
Load Data into R
```
```{r}
estate <- read_csv("../../data/estate.csv")
```

```{r}
head(estate)
```

```{block}
I'm going to change the obvious variables that should be factors to factors.
```

```{r}
estate %>%
  mutate(AC = recode(AC, "1" = "AC", "0" = "noAC"),
         Pool = recode(Pool, "1" = "Pool", "0" = "noPool"),
         Style = as.factor(Style),
         Highway = recode(Highway, "1" = "Highway", "0" = "noHighway")) ->
  estate
```


```{block}
## Univariate Distributions
```

```{r}
ggplot(estate, aes(x = Price)) +
  geom_histogram(bins = 30)

## Two bumps in Area followed by a long tail
ggplot(estate, aes(x = Area)) +
  geom_histogram(bins = 40) 

ggplot(estate, aes(x = Year)) +
  geom_histogram(bins = 20)

ggplot(estate, aes(x = Lot)) +
  geom_histogram(bins = 30)
```

```{r}
## mostly 3 and 4 bedroom houses, but there is a 0 bedroom house
## Is that a studio?
table(estate$Bed) 

## A 0 bathroom house??? Is that the same house?
table(estate$Bath)

## Let's look at that unit
estate %>%
  filter(Bath == 0)

## It's price is on the high end for having no bathroom!
## (92nd percentile)
estate %>%
  filter(Bath == 0) %>%
  select(Price) %>%
  c() ->
  weird_house_price
mean(estate$Price < weird_house_price)

## I woul dkeep in mind removing that house if I was to go on and do a 
## Linear regression
```

```{r}
table(estate$AC)

## One garage holds 7 cars?
table(estate$Garage)

table(estate$Pool)

table(estate$Quality)

table(estate$Style)

## Very few houses on a highway. I would be 
## Careful about inferences there
table(estate$Highway)
```

```{block}
## Look at bivariate associations
```

```{r, message=FALSE}
estate %>%
  select(Price, Area, Year, Lot) %>%
  ggpairs()
```

```{block}
It seems that we can log price and area and lot pretty safely.
```

```{r, message=FALSE}
estate %>%
  mutate(logPrice = log2(Price), logArea = log2(Area), logLot = log2(Lot)) ->
  estate

estate %>%
  select(logPrice, logArea, Year, logLot) %>%
  ggpairs()
```

```{block}
It seems that area has the strongest relationship to price
```

```{r}
estate %>%
  group_by(Year) %>%
  summarize(meanPrice = mean(Price)) %>%
  ggplot(aes(x = Year, y = meanPrice)) +
  geom_line()
```

```{r}
summary(estate$Year)
## Define Age of hours at sale
estate %>%
  mutate(Age = 2002 - Year) ->
  estate
```

```{block}
# Q: Does anything variable get more important with the age of the house?
```

```{r}
## Most no-ac houses are older. And once you adjust for age.
## But it still seems that there is an AC effect, which is particularly
## strong for older houses.
ggplot(estate, aes(x = Age, y = Price, color = AC)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  scale_y_log10()


## There seems to be an additive pool effect
ggplot(estate, aes(x = Age, y = Price, color = Pool)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  scale_y_log10()

## The houses near the highway almost always have a lower price
ggplot(estate, aes(x = Age, y = Price, color = Highway)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()

estate %>%
  mutate(logAge = log2(Age)) ->
  estate
```

```{block}
## Redo coloring with Area
```

```{r}
## Most no-ac houses are older. And once you adjust for age.
## But it still seems that there is an AC effect, which is particularly
## strong for older houses.
ggplot(estate, aes(x = Area, y = Price, color = AC)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  scale_y_log10()


## There seems to be an additive pool effect
ggplot(estate, aes(x = Area, y = Price, color = Pool)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  scale_y_log10()

## The houses near the highway almost always have a lower price
ggplot(estate, aes(x = Area, y = Price, color = Highway)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

```{block}
## Look at the price by those other categories
```

```{r}
estate %>% 
  ggplot(aes(x = Quality, y = Price)) +
  geom_boxplot()

## Seems that we could treat bed as a 
## Quantitative variable if we got
## rid of that 0 house
estate %>% 
  ggplot(aes(x = as.factor(Bed), y = Price)) +
  geom_boxplot() +
  scale_y_log10()

## I'm going to marge 5 bath into 5 and above,
## But still treat it as a quantitative variable
estate %>% 
  ggplot(aes(x = as.factor(Bath), y = Price)) +
  geom_boxplot() +
  scale_y_log10()

estate %>%
  mutate(Bath = recode(Bath, `6` = 5, `7` = 5)) ->
  estate

table(estate$Bath)

estate %>% 
  ggplot(aes(x = as.factor(Bath), y = Price)) +
  geom_boxplot() +
  scale_y_log10()


## Garage also looks pretty linear
estate %>%
  ggplot(aes(x = as.factor(Garage), y = Price)) +
  geom_boxplot() +
  scale_y_log10()
  
```

```{block}
## Summary of interesting Observations

- There seems to be a bathroom saturation effect. Having more than 5 doesn't 
  help you that much.
- Most quantitative relationships with price are exponential in nature. In other
  words, a multiplicative difference in area corresponds to a multiplicative
  difference in price. A multiplicative difference in age corresponds to a 
  multiplicative difference in price.
- A lot of the discrete quantities (number of garages, number of baths, 
  number of beds) can be treated as a linear relationship with log Price. So
  if you add one more bedroom, it results in some multiplicative change in 
  price.
- There is one observation with no bed, no bath, but is super pricey. My guess
  is that this is an empty lot on some prime real estate. I would exclude this
  from any model and state clearly that our analysis is for homes with at least
  one bedroom.
```




```{block}
## Some exploratory forward/backward linear model stuff
```

```{r}
estate %>%
  filter(Bath != 0) ->
  estate_sub
lmfull <- lm(logPrice ~ Bed + Bath + AC + Garage + Pool + 
               Quality + Style + Highway + logArea + logLot + logAge,
             data = estate_sub)

sout <- step(lmfull)
```

```{r}
resvec <- resid(sout)
fitvec <- fitted(sout)

## Residuals look pretty awesome
qplot(fitvec, resvec) +
  geom_hline(yintercept = 0, linetype = "dashed")

qplot(sample = resvec, geom = "qq") +
  geom_qq_line()
```

```{r}
coefvec <- coef(sout)
confintmat <- confint(sout)
sumlm <- summary(sout)
sumlm
```


```{block}
Here is some example statements:

- Multiplicative increase in price when you add one bathroom: 
  `r 2^coefvec["Bath"]` (because I used base 2 when logging), which is
  1.054 (95% CI `r 2^confintmat["Bath", ]`, which is 1.029 to 1.081).
- Doubling the area corresponds to a multiplicative increase of 
  `r 2^coefvec["logArea"]` (which is 1.693) in price (95% CI of 
  `r 2^confintmat["logArea", ]`, which is 1.58 to 1.82).
  
I used causal language here, this is a little loose and relaxed. In formal 
write ups you should use non-causal language like:

- Houses with one more bathroom tend to cost 5% more 
  (95% confidence of 3% to 8%)
- Homes with twice the area tend to cost 69% more (95% confidence of 58% to 82%).
```

```{block}
Interestingly, bed was not informative given the other variables. Why? Probably
because it is so highly correlated with the other variables that it doesn't 
add any additional information on price (at least given this dataset).
```

```{r}
ggplot(estate, aes(x = Bed, y = Bath)) +
  geom_count()
ggplot(estate, aes(x = logArea, y = logPrice, color = as.factor(Bed))) +
  geom_point()
```











