---
title: "Recoding Variable Values"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Changing the values of a variable based on its current value with `recode()`.
- Changing the values of a variable based on logical statements with `if_else()`.
- Replacing `NA`'s with `replace_na()`.

# Change variable values with `recode()`

- Variable values are sometimes uninformative. You might want to 
  change these values before plotting or calculating summary statistics.
  
- E.g. the estate data in "estate.csv" at 
  <https://dcgerard.github.io/stat_412_612/data/estate.csv>:
  
    ```{r, message = FALSE}
    library(tidyverse)
    estate <- read_csv(file = "https://dcgerard.github.io/stat_412_612/data/estate.csv")
    ```
  
- `estate` contains the following variables:
    - `Price`: Sales price of residence (in dollars)
    - `Area`: Finished area of residence (in square feet)
    - `Bed`: Total number of bedrooms in residence
    - `Bath`: Total number of bathrooms in residence
    - `AC`: `1` = presence of air conditioning, `0` = absence of air conditioning
    - `Garage`: Number of cars that a garage will hold
    - `Pool`: `1` = presence of a pool, `0` = absence of a pool
    - `Year`: Year property was originally constructed
    - `Quality`: Index for quality of construction. `High`, `Medium`, or `Low`.
    - `Style`: Categorical variable indicating architectural style 
    - `Lot`: Lot size (in square feet)
    - `Highway`: `1` = highway adjacent, `0` = highway not adjacent.  

- It would be better if we could change the `0`/`1` coding for `AC`, `Pool`,
  and `Highway` to something more informative. That way we won't have to 
  always look up the coding during our analysis.
  
- `recode()`:
    - Takes a vector as its first argument. 
    - Each subsequent argument contains two values separated by an equals sign. 
    - The value on the left of the equals sign is the current value inside the vector. 
    - The value on the right of the equals sign is the new value for the vector.
    - If the current current value inside the vector is a numeric,
      then you need to surround its value by backticks "` `` `".
    - It returns a vector with replaced values.

- Toy example:
    ```{r}
    char_vec <- c("a", "a", "b", "c", "c", "a", "b", "b", "c")
    recode(char_vec, 
           a = "Apple")
    recode(char_vec,
           b = "Banana")
    recode(char_vec,
           a = "Apple", 
           b = "Banana",
           c = "Carrot")
    ```

- **Exercise**: In the below vector, recode `"Bob"` to be `"Robert"`, `"John"` to be
  `"Jonathan"`, and `"Dave"` to be `"David"`.
  
    ```{r, echo = TRUE}
    namevec <- c("Bob", "John", "John", "John", "Dave", "Bob", "Bob", "Dave", "John")
    ```
    
    ```{r, echo = FALSE, eval = FALSE}
    recode(namevec,
           Bob = "Robert",
           John = "Jonathan",
           Dave = "David")
    ```
    
- Let's use `recode()` in to change the `Quality` values in the `estate`
  data frame. Recall: we need to use `mutate()` to modify a variable in a 
  data frame.
      
    ```{r}
    estate %>%
      mutate(Quality = recode(Quality,
                              High   = "Palace",
                              Medium = "Home",
                              Low    = "Slum")) ->
      estate
    glimpse(estate)
    ```

- Let's modify `AC`. We'll need backticks here since `1` and `0` are numerics.

    ```{r}
    estate %>%
      mutate(AC = recode(AC,
                         `1` = "AC",
                         `0` = "No AC")) ->
      estate
    glimpse(estate)
    ```

- **Exercise**: Recode the `Highway` and `Pool` variables to have more 
  informative values.
  
    ```{r, eval = FALSE, echo = FALSE}
    estate %>%
      mutate(Highway = recode(Highway,
                              `1` = "Highway",
                              `0` = "No Highway"),
             Pool = recode(Pool,
                           `1` = "Pool",
                           `0` = "No Pool"))
    ```
  
  
# Recode with Logicals with `if_else()`

- Sometimes, it is easier to recode based on logical statements.

- For example, suppose we want to recode the "Bath" variable to have values
  `1`, `2`, `3`, and `>3`. One way to do this would be:
  
    ```{r}
    estate %>%
      mutate(Bath = as.character(Bath),
             Bath = recode(Bath,
                           `4` = ">3",
                           `5` = ">3",
                           `6` = ">3",
                           `7` = ">3")) ->
      estate_temp
    glimpse(estate_temp)
    ```
    
- But this is a lot of typing. But we know how to get obtain `TRUE`'s and
  `FALSE`'s based on whether a house has more than 3 bathrooms.
  
    ```{r}
    estate %>%
      mutate(Bath > 3) %>%
      select(contains("Bath")) %>%
      glimpse()
    ```

- `if_else()`:
    - Takes a *logical* vector as its first argument.
    - It takes a vector that populates the `TRUE` values as its second argument.
    - It takes a vector that populates the `FALSE` values as its third argument.
    - The second and third arguments *must be the same type* (e.g. both logical,
      both numeric, both character, etc).
    - The second and third arguments must either be of length 1, or the same
      length as the logical vector.
    - It returns a vector with replaced values.
  
- Toy Example:

    ```{r, error = TRUE}
    x <- c(1, 2, 3, 4, 5, 6, 7, 8)
    if_else(x > 4, 4, x)
    if_else(x > 4, x, 4)
    if_else(x > 4, "x > 4", as.character(x)) 
    if_else(x > 4, "x > 4", x) ## should error 
    ```


- **Exercise**: Why did the last `if_else()` call error?
    ```{block, eval = FALSE, echo = FALSE}
    `x` is a numeric and "x > 4" is a character. `if_else()` expects the second
    and third arguments to be the same type.
    ```

- Let's apply `if_else()` to the `estate` data frame.
  Recall: we need to use `mutate()` to modify a variable in a data frame.

    ```{r}
    estate %>%
      mutate(Bath = if_else(Bath > 3,
                            ">3",
                            as.character(Bath))) ->
      estate_temp
    glimpse(estate_temp)
    ```

- **Exercise**: Recode `Price` so that any price less than \$250,000 is
  just listed as `"<250,000"`.
    ```{r, eval = FALSE, echo = FALSE}
    estate %>%
      mutate(Price = if_else(Price < 250000,
                             "<250,000",
                             as.character(Price))) %>%
      glimpse()
    ```
    
# Dealing with `NA`'s by `replace_na()`

- The `starwars` data frame from the dplyr package contains information on 
  different characters from the [Star Wars](https://en.wikipedia.org/wiki/Star_Wars)
  franchise:
    ```{r}
    data("starwars")
    glimpse(starwars)
    ```
    
- The `gender` variable is missing for some individuals
    ```{r}
    starwars %>%
      filter(is.na(gender)) %>%
      select(name, gender)
    ```
    
- But because programmers typically remove `NA`'s, it might be reasonable to change this to something else like `"other"`.
  
- `replace_na()`
    - Takes a vector as its first argument.
    - The second argument is the value with which to replace all `NA`'s.
    - It returns a vector with the `NA`'s replaced.
    
- Toy example:

    ```{r}
    x <- c("This", "is", "a", NA, NA, "vector")
    replace_na(x, "foo")
    ```

- Let's replace the `NA`'s in the `gender` variable in the `starwars` data frame.
  Recall: we need to use `mutate()` to modify a variable in a data frame.

    ```{r}
    starwars %>%
      mutate(gender = replace_na(gender, "other")) %>%
      select(name, gender) %>%
      glimpse()
    ```

- **Exercise**: In the `starwars` data frame, replace the `NA`'s in 
  `hair_color` with `"bald"`.
  
    ```{r, eval = FALSE, echo = FALSE}
    starwars %>%
      mutate(hair_color = replace_na(hair_color, "bald"))
    ```
