---
title: "R Scripts"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- R scripts
- sourcing R scripts
- Chapter 6 of [RDS](https://r4ds.had.co.nz/)

# Background

- We've already learned that using the same code over and over again in the form
  of functions is efficient and less prone to errors.
  
- What if we want to share code between multiple R Markdown files?

# R scripts

- An R Script is a file that contains just R code (no plain text like in an R 
  Markdown file).

- All R scripts end in ".R" (community standard) or ".r" (less standard).

- Open up an R script with CONTROL + SHIFT + N (or by using the File menu).

- You can then type R code in the R script and send it to the Console like in 
  an R chunk (using CONTROL/COMMAND + ENTER).
  
- Save R scripts in the "code" folder of your project.

# Using R Scripts to Reuse Code 

- Suppose we want to calculate the [geometric mean](https://dcgerard.github.io/stat_412_612/lectures/02_functions_pipes/02_functions_pipes.pdf) over and over again, in many files of our analysis.

- So far, we would copy and paste the function in every new Rmd. But we can 
  avoid this copying and pasting by coding the geometric mean function in its own R script.
  
- In the code folder, my R script "geo_mean.R" looks like this:

    ```{r, comment="", echo=FALSE}
    cat(readr::read_file("./geo_mean.R"))
    ```

- You can **source** this code (make R run the R script) by
    ```{r, eval = FALSE}
    source("../code/geo_mean.R")
    ```
    ```{r, echo = FALSE}
    source("./geo_mean.R")
    ```

- You then have access to the `geo_mean()` function in your R Markdown file:
    ```{r}
    library(magrittr)
    geo_mean(c(1, 6, 2, 5))
    ```
    
- **Exercise**: Write a function that takes a vector of numerics and returns 
  `TRUE` if that vector is between 0.1 and 0.5, and `FALSE` otherwise. Save this
  function in an R script in your code folder. Source it in an R Markdown file
  in your analysis folder. Then evaluate that function with `c(0, 0.2, 0.4, 0.7)`.



