---
title: "Dates Lab"
date: "`r Sys.Date()`"
author: "David Gerard"
output: pdf_document
urlcolor: "blue"
params: 
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$solutions, 
                      eval = params$solutions,
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Practice `lubridate`.
- More practice with `dplyr` and `tidyr`.

# Capital Bikeshare

For this exercise, we will use the data from `capital_trips_2016.csv` found
at 

<https://dcgerard.github.io/stat_412_612/data/capital_trips_2016.csv>.

1. Read these data into R.

    ```{r, message=FALSE, warning=FALSE}
    library(tidyverse)
    library(lubridate)
    ```

    ```{r, message = FALSE}
    cap <- read_csv("../../data/capital_trips_2016.csv")
    ```
    
2. Rename variables to conform to our guidelines on variable names.

    ```{r}
    cap %>%
      rename(duration             = `Duration (ms)`,
             start_time           = `Start date`,
             end_time             = `End date`,
             start_station_number = `Start station number`,
             start_station        = `Start station`,
             end_station_number   = `End station number`,
             end_station          = `End station`,
             bike_number          = `Bike number`,
             member_type          = `Member Type`) ->
      cap
    ```


3. Parse the start date and end date variables to be date-times if they 
   aren't already.

    ```{r}
    cap %>%
      mutate(start_time = mdy_hm(start_time),
             end_time   = mdy_hm(end_time)) ->
      cap
    ```
    
4. Use the start date and end date variables to calculate the duration of each trip.
   See if this duration is in agreement with the duration (ms) variable
   provided in the data set. Can you explain any discrepancies?
    ```{r}
    cap %>%
      mutate(my_duration = as.duration(end_time - start_time),
             given_duration = dmilliseconds(duration),
             dur_diff = my_duration - given_duration) ->
      cap
    
    ## Seems to be around 50 trips where the durations differ significantly.
    ggplot(cap, aes(x = dur_diff)) +
      geom_histogram(bins = 30) +
      coord_cartesian(ylim = c(0, 50))
    
    
    ## The rest of the durations differ by less than 60 seconds
    ## So this is just a result of us getting the start and end times to 
    ## the nearest minute.
    cap %>%
      filter(dur_diff < dseconds(3000)) %>%
      summarize(max_dur = max(dur_diff))
    
    ## Let's look at these weird durations
    cap %>%
      filter(dur_diff > dseconds(3000)) ->
      weird_trips
    
    ## These all occur on march 13. This is daylight savings! We run into
    ## trouble because R thinks we got these dates in UTC, not eastern.
    weird_trips %>%
      select(start_time, end_time)
    ```
    
5. If you found any discrepancies in part 3, fix them now.

    ```{r}
    cap %>%
      mutate(start_time  = force_tz(start_time, tzone = "America/New_York"),
             end_time    = force_tz(end_time, tzone = "America/New_York"),
             my_duration = as.duration(end_time - start_time),
             dur_diff    = my_duration - given_duration) ->
      cap

    cap %>%
      summarize(max_diff = max(dur_diff))
    ```

6. How much time elapsed between the start of the first trip and the end of the
   the last trip.
   
    ```{r}
    cap %>%
      summarize(min_start = min(start_time),
                max_end   = max(end_time)) ->
      extremes
    
    as.duration(extremes$max_end - extremes$min_start)
    ```
   
7. Is a trip's duration associated with the hour of day? Make a plot to explore.

    ```{r}
    cap %>%
      mutate(hour = hour(start_time)) ->
      cap
    
    ## Trips between 10AM and 5PM seem to take longer
    cap %>%
      group_by(hour) %>%
      summarize(median_duration = median(given_duration)) %>%
      ggplot(aes(x = hour, y = median_duration)) +
      geom_line()
    ```

   

 

