---
title: "Dates"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Manipulating dates and times.
- Chapter 16 of [RDS](https://r4ds.had.co.nz/).
- [Dates and Times Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/master/lubridate.pdf).
- [Lubridate Overview](https://lubridate.tidyverse.org/).

# Parsing Dates

- The lubridate package has a bunch of convenience functions for working with 
  dates. It is *not* a part of the tidyverse, so you need to load it separately.
  
    ```{r, message=FALSE, warning=FALSE}
    library(tidyverse)
    library(lubridate)
    ```
  
- There are three main classes for date/time data:
    - `Date` for just the date.
    - `POSIXct` for both the date and the time. "POSIXct" stands for 
      "Portable Operating System Interface Calendar Time" (don't ask me where 
      the "X" comes from). It is a part of a 
      [standardized system](https://en.wikipedia.org/wiki/Unix_time) 
      of representing time across many computing computing platforms.
    - `hms` from the hms R package for just the time. "hms" stands for
      "hours, minutes, and seconds."
    
- `today()` will give you the current date in the `Date` class. 
    ```{r}
    today()
    class(today())
    ```

- `now()` will give you the current date-time in the `POSIXct` class.
    ```{r}
    now()
    class(now())
    ```
    
- There is no built-in R function to find the current time without the date. 
  But you can use `hms::as.hms(now())` to get the current time.

    ```{r}
    hms::as.hms(now())
    class(hms::as.hms(now()))
    ```


# Parsing Dates

- You can use `parse_date()`, `parse_datetime()`, and `parse_time()` to 
  [parse](https://dcgerard.github.io/stat_412_612/lectures/06_tidyr/06_parsers.pdf) 
  a date/date-time/time from a string.

    ```{r}
    x <- parse_date("10/11/2020", format = "%m/%d/%Y")
    x
    class(x)
    
    y <- parse_datetime("10/11/2020 11:59:20", format = "%m/%d/%Y %H:%M:%S")
    y
    class(y)
    
    z <- parse_time("11:59:20", "%H:%M:%S")
    z
    class(z)
    ```
    
- lubridate comes with a bunch of helper functions to parse dates more 
  automatically. The helper function name itself specifies the order of the 
  year, month, day, hours, minutes, and seconds. 

- To parse dates, look at the help page of
    ```{r, eval = FALSE}
    help(ymd)
    ```

    ```{r}
    ## Only the order of year, month, and day matters
    ymd(c("2011/01-10", "2011-01/10", "20110110"))
    mdy(c("01/10/2011", "01 adsl; 10 df 2011", "January 10, 2011"))
    ```

- To parse times, look at the help page of
    ```{r}
    help(ms)
    ```
    
    ```{r}
    ## only the order of hours, minutes, and seconds matter
    hms(c("10:40:10", "10 40 10"))
    ```
    
- Note that `ms()`, `hm()`, and `hms()` won't recognize "`-`" as a separator 
  because it treats it as negative time. So use `parse_time()` here.
    ```{r}
    ms("10-10")
    ```

- To parse date-times, look at the help page of
    ```{r}
    help(ymd_hms)
    ```


- **Exercise**: Parse the following date-times.

    ```{r, eval = FALSE}
    "05/26/2004 UTC 11:11:11.444"
    "26 2004 05 UTC 11/11/11.444"
    ```
    
    ```{r, eval = FALSE, echo = FALSE}
    mdy_hms("05/26/2004 UTC 11:11:11.444")
    
    ## No dym_hms() function is defined, so need to use parse_datetime()
    parse_datetime("26 2004 05 UTC 11/11/11.444", "%d %Y %m UTC %H/%M/%S")
    ```
    
- **Exercise** (RDS 16.2.4.3): Use the appropriate lubridate function to parse 
  each of the following dates:

    ```{r}
    d1 <- "January 1, 2010"
    d2 <- "2015-Mar-07"
    d3 <- "06-Jun-2017"
    d4 <- c("August 19 (2015)", "July 1 (2015)")
    d5 <- "12/30/14" # Dec 30, 2014 
    ```

    ```{r, eval = FALSE, echo = FALSE}
    mdy(d1)
    ymd(d2)
    dmy(d3)
    mdy(d4)
    mdy(d5)
    ```
# Dates from individual components

- If you have a vector of years, months, days, hours, minutes, or seconds, you
  can use `make_date()` or `make_datetime()` to create dates and date-times.
  
    ```{r}
    make_date(year = 1981, month = 6, day = 25)
    make_datetime(year = 1972, month = 2, day = 22, hour = 10, min = 9, sec = 01)
    ```

- nycflights13 example:
    ```{r}
    library(nycflights13)
    data("flights")
    flights %>%
      mutate(datetime = make_datetime(year   = year, 
                                      month  = month, 
                                      day    = day,
                                      hour   = hour,
                                      min = minute)) ->
      flights
      select(flights, datetime)
    ```
    
- Having it in the date-time format makes it easier to plot.
    ```{r}
    ggplot(flights, aes(x = datetime)) +
      geom_freqpoly(bins = 365)
    ```

- It makes it easier to filter by date
    ```{r}
    flights %>%
      filter(as_date(datetime) == ymd(20130704)) %>%
      ggplot(aes(x = datetime)) +
      geom_freqpoly(binwidth = 600)
    ```
    
- I used `as_date()` in the previous example. This function will try to coerce
  an object to a date. Sometimes successfully! It is particularly useful for
  extracting the date component of a `POSIXct` object.

- `as_datetime()` tries to coerce an object to a `POSIXct` object.  

# Extracting Components

- `year()` extracts the year.
- `month()` extracts the month.
- `week()` extracts the week.
- `mday()` extracts the day of the month (1, 2, 3, ...).
- `wday()` extracts the day of the week (Saturday, Sunday, Monday ...).
- `yday()` extracts the day of the year (1, 2, 3, ...)
- `hour()` extracts the hour.
- `minute()` extract the minute.
- `second()` extracts the second.

    ```{r}
    ddat <- mdy_hms("01/02/1970 03:51:44")
    ddat
    year(ddat)
    month(ddat, label = TRUE)
    week(ddat)
    mday(ddat)
    wday(ddat, label = TRUE)
    yday(ddat)
    hour(ddat)
    minute(ddat)
    second(ddat)
    ```


- **Exercise**: Load the `wmata_ridership` data frame into R from
  <https://dcgerard.github.io/stat_412_612/data/wmata_ridership.csv>. For each 
  month, calculate the proportion of rides made on a given day of the month. Then
  make box plots of the proportions of ridership vs day of the weak. But exclude
  any days from 2004.
  
    ```{r, eval = FALSE, echo = FALSE}
    wmata <- read_csv("../../data/wmata_ridership.csv")
    
    wmata %>%
      mutate(year = year(Date),
             month = month(Date),
             day  = wday(Date, label = TRUE)) %>%
      filter(year != 2004) %>%
      group_by(year, month) %>%
      mutate(prop = Total / sum(Total)) %>%
      ungroup() ->
      wmata
    
    ggplot(wmata, aes(x = day, y = prop)) +
      geom_boxplot()
    ```
  
- You can overwrite components.

    ```{r}
    ddat <- mdy_hms("01/02/1970 03:51:44")
    ddat
    year(ddat) <- 1988
    ddat
    ```
    
- You can round components

    ```{r}
    ddat <- mdy_hms("01/02/1970 03:51:44")
    ddat
    round_date(ddat, unit = "year")
    ```

  
# Time Spans

- To count the number of seconds between two dates, use a `duration`. You can
  read about durations using
    ```{r, eval = FALSE}
    help("Duration-class")
    ```

- We can find out how old 
  [Patrick Stewart](https://en.wikipedia.org/wiki/Patrick_Stewart) 
  is using durations
    ```{r}
    d1 <- ymd(19400713)
    d2 <- today()
    agesec <- as.duration(d2 - d1)
    agesec
    ```

- You can create durations from years with `dyears()`, from days with 
  `ddays()`, etc...
  
    ```{r}
    dyears(1)
    ddays(1)
    dhours(1)
    dminutes(1)
    dseconds(1)
    ```
  


- You can add durations to date-times, but you always add *seconds*, so if there
  is daylight savings you get weird results (add a day but the time is not the same
  as the time the previous day).
  
    ```{r}
    one_pm <- ymd_hms("2016-03-12 13:00:00", tz = "America/New_York")
    one_pm
    one_pm + ddays(1)
    ```
    
- Adding a `period` takes into account daylight savings.
    ```{r}
    one_pm
    one_pm + days(1)
    ```

- You can read more about periods with
    ```{r, eval = FALSE}
    help("Period-class")
    ```

- Intervals are like durations, but they also have an associated start time. You
  can read more about intervals with
    ```{r, eval = FALSE}
    help("Interval-class")
    ```
    
- **Exercise**: How long of a time-span is covered in the WMATA ridership 
  dataset?
  
    ```{r, eval = FALSE, echo = FALSE}
    wmata <- read_csv("../../data/wmata_ridership.csv")
    rng <- range(wmata$Date)
    as.duration(rng[2] - rng[1])
    as.interval(rng[2] - rng[1], start = rng[1]) / years(1)
    ```
    
# Time Zones

- Time zones are specified using the `tz` or `tzone` arguments (for example, in the call
  to `ymd_hms()` above).
  
- Time zones are specified by "content/city." For example, `"America/New_York"`
  and `"Europe_Paris"`
  
- You can see a complete list of time zones with `OlsonNames()`.

- The default time zone is `UTC` (which has no daylight savings).

- You usually don't have to worry about timezones unless you loaded them in
  incorrectly. For example, R might think it's `UTC` even though it should be 
  `America/New_York` and then forget daylight savings.
  
- If a date-time is labelled with the incorrect time zone, use `force_tz()`.

    ```{r}
    d1 <- ymd_hms("20140101 10:01:11")
    d1
    force_tz(d1, tzone = "America/New_York")
    ```

- If the timezone is correct, but you want to change it, use `with_tz()`.
  
    ```{r}
    with_tz(d1, tzone = "America/New_York")
    ```

