---
title: "Factors Lab"
date: "`r Sys.Date()`"
author: "David Gerard"
output: pdf_document
urlcolor: "blue"
bibliography: "fac_bib.bib"
params: 
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$solutions, eval = params$solutions)
```

# Learning Objectives

- Practice `forcats`

# Exercise 1: Factor Coding Exercises

Consider the factor vector

```{r, echo = TRUE}
fc <- factor(c("D", "C", "C", "A", "A", 
               "C", "D", "A", "C", "C", 
               "A", "A", "C", "A", "C", 
               "B", "C", "A", "C", "B"))
```

1. `fc` should also have a level called `E`. Add this level to `fc`.

```{r}
fct_expand(fc, "E")
```

2. Plot the distribution of `fc`, ordering levels from left to right by 
   frequency (so most frequent is on the right).

    ```{r}
    tibble(fc = fc) %>%
      mutate(fc = fct_rev(fct_infreq(fc))) %>%
      ggplot(aes(x = fc)) +
      geom_bar()
    ```
    
3. Put the levels in *reverse* alphabetical order.
    ```{r}
    fct_relevel(fc, sort(levels(fc), decreasing = TRUE))
    ```
    
4. Remove all `"C"`'s from `fc` then drop this level.

    ```{r}
    tibble(fc = fc) %>%
      filter(fc != "C") %>%
      mutate(fc = fct_drop(fc)) %>%
      count(fc)
    ```

5. Combine the `"A"` and `"B"` levels into the new level `"AB"`.

    ```{r}
    fct_collapse(fc, "AB" = c("A", "B"))
    ```



# Exercise 2: Sex, Lies, and Religion

The data frame in <https://dcgerard.github.io/stat_412_612/data/sexlierel.txt>,
taken from @clayton1971religiosity, contain the following variables:

- `gender`: The gender of the individual. `1` = Female, `2` = Male.
- `scale`: Different scales of sexual permissiveness. `1` = Ritualistic, `2` = Experiential, `3` = Ideological, `4` = Composite.
- `perm`: The level of sexual permissiveness. `1` = Low, `2` = High.
- `lie`: The propensity to lie. `1` = Lower, `2` = Higher.
- `relig`: How religious a person is. `1` = Low, `2` = High
- `count`: The number of individuals satisfying the conditions of the other variables.


1. Read the data into R.

    ```{r}
    library(tidyverse)
    sexlierel <- read_table(file = "../../data/sexlierel.txt", 
                            col_types = cols(
                              gender = col_factor(),
                              scale  = col_factor(),
                              perm   = col_factor(),
                              lie    = col_factor(),
                              relig  = col_factor(),
                              count  = col_double()
                              ))
    ```

2. Change the level names to something more informative.

    ```{r}
    sexlierel %>%
      mutate(gender = fct_recode(gender, 
                                 "Female" = "1", 
                                 "Male"   = "2"),
             scale  = fct_recode(scale,
                                 "ritualistic"  = "1", 
                                 "experiential" = "2",
                                 "idiological"  = "3",
                                 "composite"    = "4"),
             perm   = fct_recode(perm,
                                 "low"  = "1",
                                 "high" = "2"),
             lie    = fct_recode(lie,
                                 "lower"  = "1",
                                 "higher" = "2"),
             relig  = fct_recode(relig,
                                 "low"  = "1",
                                 "high" = "2")) ->
      sexlierel
    sexlierel
    ```
    
3. Flip the order of the levels in `perm`.
    ```{r}
    sexlierel %>%
      mutate(perm = fct_rev(perm)) ->
      sexlierel
    ```

    
4. For males, for what scales does there appear to be an association between 
   religiosity and permissiveness? Use one plot to explore.
  
    ```{r}
    ## All of them
    sexlierel %>%
      filter(gender == "Male") %>%
      ggplot(aes(x = relig, y = count, fill = perm)) +
      geom_col(pos = "fill") +
      facet_grid(. ~ scale) +
      ylab("Proportion") +
      xlab("Religiosity") +
      scale_fill_discrete(name = "Permissiveness") +
      theme(strip.background = element_rect(fill = "white"))
    ```






# References
