---
title: "Factors"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Manipulating factors.
- Chapter 15 of [RDS](https://r4ds.had.co.nz/).
- [Wrangling Categorical Data in R](https://doi.org/10.7287/peerj.preprints.3163v2).
- [8.2: Chimeras of the R Inferno](https://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
- [Factors with forcats Cheat Sheet](https://www.rstudio.com/resources/cheatsheets/#forcats).

# Factors

- A "factor" is R's way to say that a variable is categorical 
  (places observational/experimental units into different groups or categories 
  based on its values.).
  
- A factor is different from a character in that:
    1. There is a small predefined set of "levels" (possible values) of a 
       factor, but not of a character.
    2. There is an ordering for the levels of a factor 
        - Useful when determining the order to plot something.
        - Useful when doing orderd logistic regression.
        
- Consider the following data frame for average highs in DC for each month.

    ```{r, message = FALSE}
    library(tidyverse)
    dcclimate <- tribble(~month, ~avehigh,
                         ##----/---------
                         "Jan",  43.4,
                         "Feb",  47.1,
                         "Mar",  55.9,
                         "Apr",  66.6,
                         "May",  75.4,
                         "Jul",  88.4,
                         "Aux",  86.5,
                         "Sep",  79.5,
                         "Oct",  68.4,
                         "Nov",  57.9,
                         "Dec",  46.8)
    ```
    
- The weather for June is missing and the 3-letter abbreviation for August is 
  incorrect. We would like to notice both of these.
  
- Also, when we plot the data, we would prefer the order to be the same as that
  for the order of the months of the year.
  
    ```{r}
    ggplot(dcclimate, aes(x = month, y = avehigh)) +
      geom_col()
    ```
    
- Factors help us with all of these issues.

- You have to be **very** careful about factors.

    ```{r}
    x  <- c("51", "32", "15", "2", "32")
    xf <- factor(x)
    as.numeric(x)
    as.numeric(xf)
    
    as.numeric("Hello")
    as.numeric(factor("Hello"))
    
    fac1 <- factor(c("x1", "x2", "x3"))
    fac2 <- factor(c("y1", "y2", "y3"))
    c(fac1, fac2)
    ```


# Creating Factors

- Use `factor()` or `parse_factor()` to create a factor variable 

- `parse_factor()` returns better warnings, so I would recommend always using
  that.

    ```{r}
    monthvec <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    dcclimate %>%
      mutate(monthfc = factor(month, levels = monthvec)) ->
      dcclimate
    
    dcclimate %>%
      mutate(monthfc2 = parse_factor(month, levels = monthvec)) ->
      dcclimate
    
    dcclimate$monthfc
    ```
    
- If you do not specefy the `levels` argument, R will assume that the levels
  are the unique values of the vector
    - `factor()` takes the order of the levels to be the same order returned 
      by `sort()`.
    - `parse_factor()` takes the order of the levels to be the same order
      as the order of the value introduced.
  
  
    ```{r}
    x <- c("A", "string", "vector", "is", "a", "string", "vector")
    factor(x)
    sort(unique(x))
    parse_factor(x)
    ```

- You can always see the levels of a factor (and their order) using 
  the `levels()` function
    ```{r}
    levels(dcclimate$monthfc)
    ```
- More elegant is the `count()` function:
    ```{r}
    dcclimate %>%
      count(monthfc)
    ```

- Once we have a factor variable, the order of the aesthetic map is set in ggplot.

    ```{r}
    ggplot(dcclimate, aes(x = monthfc, y = avehigh)) +
      geom_col()
    ```
    
- We can include missing levels by using the `drop = FALSE` argument in the 
  appropriate scale call:
  
    ```{r}
    ggplot(dcclimate, aes(x = monthfc, y = avehigh)) +
      geom_col() +
      scale_x_discrete(drop = FALSE)
    ```

# forcats

- forcats is an R package which makes two things much easier in R:
    - Changing the order of the levels of the factor variable.
    - Changing the levels of the factor variable.
  
- forcats is a part of the tidyverse, so you don't need to load it separately
  when you load the tidyverse.
    

## Changing the Order of the Levels

- Consider the subset of the 
  [General Social Survey](https://en.wikipedia.org/wiki/General_Social_Survey)
  stored in the `gss_cat` data in forcats.
    ```{r}
    data(gss_cat)
    glimpse(gss_cat)
    ```

- You often want to change the order of the levels of a factor to make 
  plots more insightful.
  
    ```{r}
    gss_cat %>%
      group_by(relig) %>%
      summarize(tvhours_mean = mean(tvhours, na.rm = TRUE)) ->
      tvdat
    
    ggplot(tvdat, aes(x = tvhours_mean, y = relig)) +
      geom_point() +
      xlab("Average TV Hours") +
      ylab("Religion")
    ```
  






