---
title: "Databases and dbplyr"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Using dplyr-like syntax for databases.
- [Introduction to dbplyr](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html)

# dbplyr

- [SQL](https://en.wikipedia.org/wiki/SQL) is a language used to query from
  relational datasets.

- dplyr basically implements the most common actions in SQL (but SQL can do more).

- We'll use a soccer dataset to demonstrate how to use dplyr syntax when 
  interacting with a database. Download and unzip the soccer database from
  <https://dcgerard.github.io/stat_412_612/data.html>.
  
- We'll use the dbplyr package to interact with databases.

    ```{r, eval = FALSE}
    install.packages("dbplyr")
    ```
    ```{r, message = FALSE}
    library(tidyverse)
    library(dbplyr)
    ```
- dbplyr allows you to work with databases as if you are using dplyr.

- You'll also need to install the RSQLite package. There are different ways to
  create/access/update/delete data from relational databases, and RSQLite 
  provides an R interface for one of these ways.
  
    ```{r, eval = FALSE}
    install.packages("RSQLite")
    ```
    ```{r}
    library(RSQLite)
    ```
    
- If your database uses a different engine, you'll need to download other
  packages to interact with it (see [Introduction to dbplyr](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html))
    
- First, we'll tell R where the database is using `dbConnect()`,
  (you might need to change the path).

    ```{r}
    con <- dbConnect(drv = SQLite(), dbname = "../../data/soccer/soccer.sqlite")
    ```

- Now we'll list the data frames available in the connection we just created.
    ```{r}
    dbListTables(con)
    ```
    
- Use `tbl()` to make a reference to the tables in `con`.

    ```{r}
    Team_db    <- tbl(con, "Team")
    Team_at_db <- tbl(con, "Team_Attributes")
    Country_db <- tbl(con, "Country")
    League_db  <- tbl(con, "League")
    Match_db   <- tbl(con, "Match")
    ```
    
- We can now interact with all of these data frames mostly like if they
  were in memory (with some limitations).

    ```{r}
    head(Country_db)
    head(Match_db)
    Match_db %>%
      select(id:away_team_goal)
    
    names(Match_db) ## won't work
    ```
    
- Once you select the variables you want and the observations you want, you 
  should use `collect()` to get the data frame into memory so that you can have
  all of the functionality of R (e.g., `gather()` and `spread()` will only
  work on in-memory data frames).
  
    ```{r}
    Match_db %>%
      select(id:away_team_goal) %>%
      collect() ->
      Match
    Team_db %>%
      collect() ->
      Team
    Country_db %>%
      collect() ->
      Country
    ```
    
- The following will return a data frame telling you where each country is from.
    
    ```{r}
    Match %>%
      select(country_id, home_team_api_id, away_team_api_id) %>%
      gather(-country_id, key = "home_away", value = "team_api_id") %>%
      select(-home_away) %>%
      distinct() %>%
      left_join(Team, by = "team_api_id") %>%
      left_join(Country, by = c("country_id" = "id")) %>%
      select(team_long_name, team_short_name, name) %>%
      rename(country_name = name)
    ```
    
- **Exercise**: Extract all matches from the `England Premier League` and 
  calculate the mean team difference (average of home team goals minus away
  team goals) each day in the `"2010/2011"` season. Plot this proportion 
  against time. (hint: you'll need separate date and time. You'll also need to 
  use before you plot `parse_date()`).
  
    Your plot should look like this:
  
    ```{r, echo = FALSE}
    Match_db %>%
      select(league_id, season, date, home_team_goal, away_team_goal) %>%
      filter(league_id == 1729, season == "2010/2011") %>%
      collect() ->
      subMatch
    
    subMatch %>%
      separate(col = "date", into = c("date", "time"), sep = " ") %>%
      select(-time) %>%
      group_by(date) %>%
      summarize(mean_diff = mean(home_team_goal - away_team_goal)) %>%
      mutate(date = parse_date(x = date, format = "%Y-%m-%d")) ->
      ave_diff
    
    ggplot(data = ave_diff, mapping = aes(x = date, y = mean_diff)) +
      geom_line() +
      xlab("Date") +
      ylab("Mean Difference") +
      theme_bw()
    ```
  
    
  






