---
title: "Relational Data Lab"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
params: 
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$solutions, eval = params$solutions)
```

We'll work more with the nycflights13 dataset here.

1. Load the nycflights13 data frames into R.

    ```{r}
    library(tidyverse)
    library(nycflights13)
    data("airlines")
    data("airports")
    data("flights")
    data("planes")
    data("weather")
    ```
    
2. Explore if the average arrival delay is associated with the time zone of the 
   destination.

    ```{r}
    ## Yes, points in time zones further from NYC time zone have on average
    ## longer arrival delays. This is probably just a result of distance from NYC.
    flights %>%
      group_by(dest) %>%
      summarize(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
      left_join(airports, by = c("dest" = "faa")) %>%
      select(arr_delay, tz) %>%
      mutate(tz_fac = factor(tz)) %>%
      ggplot(aes(x = tz_fac, y = arr_delay)) +
      geom_boxplot() +
      theme_bw() +
      xlab("Time Zone") +
      ylab("Mean Arrival Delay")
    ```
    
3. Calculate the average longitude and latitude of the three NCY airports. 
   We will use this as the location of NYC.
   
    ```{r}
    flights %>%
      select(origin) %>%
      distinct() %>%
      left_join(airports, by = c("origin" = "faa")) %>%
      summarize(lon = mean(lon), lat = mean(lat)) ->
      nyc_location
    nyc_location
    ```
    
4. Create a function that takes two vectors (`lon` and `lat`) and two scalars
   (`nyclon` and `nyclat`) as input. It then uses the `distm()` function in the
   geosphere R package to calculate the geodesic distance between each each 
   element in `lon` and `lat` and the location defined by `nyclon` and `nyclat`.
   For example, in my implementation:
   
    ```{r, eval = TRUE}
    library(geosphere)
    dist_nyc <- function(lon, lat, nyclon, nyclat) {
      c(distm(cbind(lon, lat), cbind(nyclon, nyclat)))
    }
    ```

    ```{r, eval = TRUE, echo = TRUE}
    lat    <- c(41.1, 32.5, 42.0)
    lon    <- c(-80.6, -85.7, -88.1)
    nyclon <- -73.9
    nyclat <- 40.8
    dist_nyc(lon = lon, lat = lat, nyclon = nyclon, nyclat = nyclat)
    ```

   
5. Use the function you created in part 4 and the NYC location you calculated
   in part 3 to find the geodesic distance from NYC of every airport.
   Explore if average arrival delay is associated with distance of the
   destination from NYC.
   
    ```{r}
    airports %>%
      mutate(distNYC = dist_nyc(lon = lon, 
                                lat = lat, 
                                nyclon = nyc_location$lon,
                                nyclat = nyc_location$lat)) ->
      airports
    
    ## Seems that delay increases then decreases with distance
    flights %>%
      group_by(dest) %>%
      summarize(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
      left_join(airports, by = c("dest" = "faa")) %>%
      ggplot(aes(x = distNYC, y = arr_delay)) +
      geom_point() +
      geom_smooth(se = FALSE) +
      theme_bw() +
      xlab("Distance from NYC") +
      ylab("Average Arrival Delay")
    ```
   
6. For each plane calculate the proportion of times that it takes off in fair 
   weather and see if that proportion is associated with the age of the plane.
   We'll define "fair weather" to mean no precipitation, wind-speeds of
   under 20 mph, and a visibility of at least 10 miles.
   
    ```{r}
    ## looks like there is no association
    flights %>%
      select(tailnum, origin, year, month, day, dep_time) %>%
      mutate(hour = dep_time %/% 100) %>%
      left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
      mutate(is_fair = precip == 0 & wind_speed < 20 & visib >= 10) %>%
      group_by(tailnum) %>%
      summarize(prop_fair = mean(is_fair, na.rm = TRUE)) %>%
      left_join(planes, by = "tailnum") %>%
      mutate(age = 2013 - year) %>%
      ggplot(aes(x = age, y = prop_fair)) +
      geom_point() +
      theme_bw() +
      xlab("Age") +
      ylab("Proportion Fair") +
      geom_smooth(method = "lm", se = FALSE)
    ```
   



