---
title: "Base R Graphics Cheat Sheet"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 3, fig.align = "center")
```

# Abstract:
I reproduce some of the plots from Rstudio's [ggplot2](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) cheat sheet using Base R graphics. I didn't try to pretty up these plots, but **you should**.

```{r, echo=FALSE}
set.seed(1)
```

I use this dataset
```{r}
data(mpg, package = "ggplot2")
```

# General Considerations

The main functions that I generally use for plotting are

- Plotting Functions
    - `plot`: Makes scatterplots, line plots, among other plots.
    - `lines`: Adds lines to an already-made plot.
    - `par`: Change plotting options.
    - `hist`: Makes a histogram.
    - `boxplot`: Makes a boxplot.
    - `text`: Adds text to an already-made plot.
    - `legend`: Adds a legend to an already-made plot.
    - `mosaicplot`: Makes a mosaic plot.
    - `barplot`: Makes a bar plot.
    - `jitter`: Adds a small value to data (so points don't overlap on a plot).
    - `rug`: Adds a rugplot to an already-made plot.
    - `polygon`: Adds a shape to an already-made plot.
    - `points`: Adds a scatterplot to an already-made plot.
    - `mtext`: Adds text on the edges of an already-made plot.
- Sometimes needed to transform data (or make new data) to make appropriate plots:
    - `table`: Builds frequency and two-way tables.
    - `density`: Calculates the density.
    - `loess`: Calculates a smooth line.
    - `predict`: Predicts new values based on a model.

All of the plotting functions have arguments that control the way the plot looks. You should read about these arguments. In particular, read carefully the help page `?plot.default`. Useful ones are:

- `main`: This controls the title.
- `xlab`, `ylab`: These control the x and y axis labels.
- `col`: This will control the color of the lines/points/areas.
- `cex`: This will control the size of points.
- `pch`: The type of point (circle, dot, triangle, etc...)
- `lwd`: Line width.
- `lty`: Line type (solid, dashed, dotted, etc...).

# One Variable

## Continuous

Density plot
```{r}
density_object <- density(mpg$hwy)
plot(density_object)
```

Histogram
```{r}
hist(mpg$hwy, breaks = 10)
```

\vspace{20mm}

## Discrete
Barplot
```{r}
tabdat <- table(mpg$drv)
barplot(tabdat)
```

Different type of bar plot
```{r}
plot(tabdat)
```

\vspace{30mm}

# Two Variables

## Continuous X, Continuous Y

Scatterplot
```{r}
plot(mpg$cty, mpg$hwy)
```

Jitter points to account for overlaying points.
```{r}
x <- jitter(mpg$cty)
y <- jitter(mpg$hwy)
plot(x, y)
```

\vspace{30mm}

Add a rug plot
```{r}
plot(x, y)
rug(x, side = 1)
rug(y, side = 2)
```

Add a Loess Smoother
```{r}
loess_fit <- loess(hwy ~ cty, data = mpg)
xnew <- seq(min(x), max(x), length = 100)
ynew <- predict(object = loess_fit, newdata = data.frame(cty = xnew))
plot(x, y)
lines(xnew, ynew, col = 2, lty = 2)
```

\vspace{20mm}

Loess smoother with upper and lower 95\% confidence bands
```{r, fig.height=4, fig.width=6}
loess_fit <- loess(hwy ~ cty, data = mpg)
xnew <- seq(min(x), max(x), length = 100)
pfit <- predict(object = loess_fit, newdata = data.frame(cty = xnew), se = TRUE)
ynew <- pfit$fit
upper_bound <- pfit$fit + qnorm(0.975) * pfit$se.fit
lower_bound <- pfit$fit - qnorm(0.975) * pfit$se.fit
plot(x, y)
lines(xnew, ynew, col = 2, lty = 2)
lines(xnew, upper_bound, col = 4, lty = 2)
lines(xnew, lower_bound, col = 4, lty = 2)
```


\vspace{70mm}

Loess smoother with upper and lower 95\% confidence bands and that fancy shading from `ggplot2`.
```{r, fig.height=4, fig.width=6}
loess_fit <- loess(hwy ~ cty, data = mpg)
xnew <- seq(min(x), max(x), length = 100)
pfit <- predict(object = loess_fit, newdata = data.frame(cty = xnew), se = TRUE)
ynew <- pfit$fit
upper_bound <- pfit$fit + qnorm(0.975) * pfit$se.fit
lower_bound <- pfit$fit - qnorm(0.975) * pfit$se.fit
xshade <- c(xnew, xnew[length(xnew):1])
yshade <- c(upper_bound, lower_bound[length(lower_bound):1])
plot(x, y)
lines(xnew, ynew, col = 2, lty = 2)
polygon(xshade, yshade, col = "#0000FF33", border = FALSE)
```


\vspace{70mm}

Add text to a plot
```{r}
plot(x, y)
text(15, 40, "some text")
```


## Discrete X, Continuous Y

Boxplot
```{r}
boxplot(hwy ~ class, data = mpg)
```


\vspace{20mm}

## Discrete X, Discrete Y

Mosaic Plot
```{r}
tabdat <- table(mpg$fl, mpg$drv)
mosaicplot(tabdat, color = TRUE)
```

## Continuous Function

Line plot
```{r}
f <- function(x) {
  return(x ^ 2)
}

x <- seq(-2, 2, length = 100)
y <- f(x)
plot(x, y, type = "l")
```

# Color Coding and Legends
Color code a scatterplot by a categorical variable and add a legend.
```{r}
x <- jitter(mpg$hwy)
y <- jitter(mpg$cty)
z <- factor(mpg$drv)
plot(x, y, col = z)
legend("topleft", legend = levels(z), col = 1:nlevels(z), pch = 1, title = "Drive")
```


\vspace{100mm}

# Faceting

`par` sets the graphics options, where `mfrow` is the parameter controling the facets.
```{r, fig.width=6, fig.height=4}
old_options <- par(mfrow = c(2, 3))
plot(mpg$cty, mpg$hwy)
hist(mpg$cty)
plot(density(mpg$cty))
plot(table(mpg$fl))
barplot(table(mpg$fl))
plot(table(mpg$fl, mpg$drv))
par(old_options)
```

The first line sets the new options and saves the old options in the list `old_options`. The last line reinstates the old options.




