---
title: "Graphics with ggplot2"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- ggplot2.
- Basics of data visualization.
- Chapter 3 [RDS](https://r4ds.had.co.nz/).
- [Data Visualization Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf).
- [Ggplot2 Overview](https://ggplot2.tidyverse.org/).

# Background

- The three main plotting systems in R are [base R](https://dcgerard.github.io/stat_412_612/lectures/03_graphics/03_base_r_cheatsheet.pdf), [lattice](https://dcgerard.github.io/stat_412_612/lectures/03_graphics/03_lattice_cheatsheet.pdf), and [ggplot2](https://www.rstudio.com/resources/cheatsheets/#ggplot2) (from the tidyverse) (with base R-like syntax implemented in the [`qplot()` function](https://dcgerard.github.io/stat_412_612/lectures/03_graphics/03_qplot_cheatsheet.pdf)).

- In base R: start with a blank plot and build on it by adding elements --- the "artist's canvas".
- In lattice: One function, specify everything at once.
- ggplot2: A "grammar of graphics". You map from variables to aesthetic attributes (size, color, shape) of geometric objects (lines, bars, points).

- Let's look at the three ways to make a scatterplot with a loess smoother and a 
  rug plot.

    - Load data and packages:
    ```{r}
    library(ggplot2)
    library(lattice)
    data("mpg")
    ```
    
    
    \newpage
    
    - base R: Everything is a separate non-connected function
    ```{r, fig.width = 6, fig.height = 4}
    ## Make a scatterplot
    plot(x = mpg$displ, y = mpg$hwy, xlab = "Displacement", ylab = "Highway")
    
    ## Add loess curve
    lout <- loess(hwy ~ displ, data = mpg)
    ord_x <- order(lout$x)
    lines(lout$x[ord_x], lout$fitted[ord_x])
    
    ## Add two rug plots
    rug(mpg$displ, side = 1)
    rug(mpg$hwy, side = 2)
    ```
    
    \newpage
    
    - lattice: Everything is done at once within some mega-function:
    ```{r, fig.width = 6, fig.height = 4}
    xyplot(hwy ~ displ, data = mpg,
           xlab = "Displacement", ylab = "Highway",
           panel = function(x, y) {
             panel.xyplot(x = x, y = y)                   ## Add scatterplot
             panel.loess(x = x, y = y, span = 0.75, 
                         family = "gaussian", degree = 2) ## Add loess curve
             panel.rug(x = x, y = y)                      ## Add rug plot
           })
    ```
    
    \newpage
    
    - ggplot2: Many connected functions
    ```{r, fig.width = 6, fig.height = 4}  
    ggplot(mpg, mapping = aes(x = displ, y = hwy)) + ## set aesthetic maps
      geom_point() +                                 ## apply aesthetic maps to 
                                                     ##     point objects
      geom_smooth(se = FALSE, method = loess) +      ## apply to smooth objects
      geom_rug() +                                   ## apply to rug objects
      xlab("Displacement") +
      ylab("Highway")
    ```
    

- lattice is the hardest to learn and the least used.
- base R is very flexible, but harder to use for most tasks.
- ggplot2 is very easy to use for most tasks, but very hard to 
  use when you want to do something weird.
  
# ggplot2 Foundation

- ggplot consists of the following elements:

    - **Data Frame** (`data`): Contains the variables that you want to plot.
    - Transformation (`stat`): Transforms data into a new data frame (e.g. by
      calculating the number of observational units in each bin for a 
      histogram).
    - **Aesthetic Mapping** (`mapping`): Assign variables to be aesthetics (e.g. 
      $x$-coordinate, $y$-coordinate, color, shape, transparency, line 
      thickness, line type) of geometric objects.
    - **Geometric Objects** (`geom`): The type of plot you want (e.g. scatterplot,
      line plot, histogram, barplot)
    - Position Adjustment (`position`): Jitter objects, overlap histogram bins, 
      etc..
    - **Facets**: Create multiple plots, one for each value of a categorical
      variable.
    - Coordinate systems: Cartesian/polar
    
- You can get most of your plotting done by understanding the basics of data
  frames, aesthetic mappings, geoms, and facets (the other aspects of the 
  grammar of graphics are less often useful to modify).

- `ggplot()` creates a blank `gg` objects. It has two main arguments:

    - `data`: Must be a data frame. Contains the variables that you want to 
      map onto aesthetics.
    - `mapping`: A definition for which variables map onto which aesthetics. 
      You almost always use the `aes()` function to perform this mapping.
      
      
    - What is the output of the following?  
    ```{r, eval = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy))
    ```


- `ggplot()` is almost always followed by a `+` and then a `geom` argument to 
  specify which geometric objects receive the aesthetic mapping. We will 
  now go through many different types of geoms and the situations where they 
  are useful:
  
- **Always make sure the `+` is at the end of a line and not at the beginning of a line.**

- To get a list of the possible aesthetic mappings of a geom, look at the help
  page.
  
# Comparing Two Quantitative Variables

- A scatterplot is used to explore the relationship between two **quantitative**
  variables. 
  
- "Quantitative" means that arithmetic operations ($+/-/\times/\div$) make sense
  on them.
  
- E.g. `hwy` and `displ` are two quantitative variables.

- Not all numeric variables are quantitative (e.g. phone numbers).

- Use `geom_point()` to create a scatterplot:

    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point()
    ```

- `displ` is mapped onto the `x` aesthetic and `hwy` is mapped onto the 
  `y` aesthetic.
  
- The above indicates a **negative** relationship between `hwy` and `displ`

- The following indicates a **positive** relationship between `hwy` and `cty`

    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
          geom_point()
    ```
    
- **Exercise**: Create a scatterplot of engine displacement vs city miles per gallon. Think
  carefully about which variable should be on the $x$-axis.
  
    ```{r, eval = FALSE, echo = FALSE}
    ggplot(mpg, aes(x = displ, y = cty)) +
      geom_point()
    ```
  

  
## Annotating by a categorical variable

- We might be interested in the association of two quantitative variables
  at different values of a categorical variable.
  
- A **categorical** variable places observational units into different groups
  or categories based on the level of the categorical variable.
  
- We often map categorical variables to the `color` (better) or `shape` 
  (worse) aesthetics:

    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
          geom_point()
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, shape = drv)) +
          geom_point()
    ```

- **Exercise**: What happens when you map a categorical variable to the `size`
  or `alpha` aesthetics?
  
```{r, eval = FALSE, echo = FALSE}
## There is no real ordering of the values of hte categorical variable, so
## the alpha and size mappings don't make sense (because there is a clear 
## ordering for these aesthetics).
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, size = drv)) +
  geom_point()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, alpha = drv)) +
  geom_point()
```
  
  
- **Exercise**: Consider the `iris` dataset that comes with R by default:
    ```{r}
    data("iris")
    ```
    
    Make a plot of petal length vs petal width, color coding by the species of 
    iris. What do you notice?
    
    ```{r, eval = FALSE, echo = FALSE}
    ## Good separation of species using two variables.
    ggplot(data = iris, mapping = aes(x = Petal.Length, 
                                      y = Petal.Width, 
                                      color = Species)) +
      geom_point()
    ```
    


    

## Annotating by a quantitative variable

- If we want to look at the association between two quantitative variables at
  different levels of a third quantitative variable, we can map that third
  quantitative variable to the `color` (good), `alpha` (good), or size (worst)
  aesthetic.
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = cty)) +
              geom_point()
    ```

    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, alpha = cty)) +
              geom_point()
    ```

    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, size = cty)) +
                  geom_point()
    ```

- **Exercise**: What happens when we map a quantitative variable to the `shape`
  aesthetic?
  
    ```{r, echo = FALSE, eval = FALSE}
    ## We get an error
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, shape = cty)) +
                  geom_point()
    ```
  
    
- We can map multiple variables to multiple aesthetics at the same time, but 
  the resulting plot might get too complicated.
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, 
                                     y = hwy,
                                     alpha = cty, 
                                     shape = drv,
                                     color = class)) +
              geom_point()
    ```

- **Exercise**: Using the `iris` dataset, plot petal length vs petal width,
  annotating by sepal length and sepal width and species. Do you notice anything?
  
    ```{r, echo = FALSE, eval = FALSE}
    ## Virginica appears to have larger sepal length. But the plot is too cluttered
    ## for me to make anything else out.
    ggplot(data = iris, mapping = aes(x = Petal.Length,
                                      y = Petal.Width,
                                      color = Species,
                                      size = Sepal.Length,
                                      alpha = Sepal.Width)) +
      geom_point()
    ```
  
  
## Smoothing

- `geom_smooth()` will create trend lines (either non-parametric or parametric)
  and plot these.
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
              geom_smooth(se = FALSE)
    ```
    
- You can pass categorical variables to the `color` (best) or `linetype` (good)
  aesthetics
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
              geom_smooth(se = FALSE)
    ```
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, linetype = drv)) +
              geom_smooth(se = FALSE)
    ```

- The default for `geom_smooth()` is to calculate a loess smoother, but other 
  options are available. For example, we can plot the OLS fit by changing the 
  `method` argument.
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
              geom_smooth(se = FALSE, method = lm)
    ```
  
- If you have binary response data, you can plot the fitted logistic curves by
    ```{r}
    ggplot(data = mtcars, mapping = aes(x = hp, y = vs)) +
      geom_smooth(method = glm, method.args = list(family = "binomial"), se = FALSE)
    ```
  
## Layering Geoms:

- You can layer multiple geoms that use the same data and aesthetic map as defined
  in the `ggplot()` call:
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point() +
      geom_smooth(se = FALSE)
    ```
  
    
- You can equivalently define the aesthetics in the particular geom you are 
  using. However, later geoms will not have access to these maps and you would
  have to redefine them.

    ```{r}
    ggplot(data = mpg) +
      geom_point(mapping = aes(x = displ, y = hwy))
    ```
    
    ```{r, error = TRUE, eval = FALSE}
    ## Should produce an error
    ggplot(data = mpg) +
      geom_point(mapping = aes(x = displ, y = hwy)) +
      geom_smooth()
    ```
  
    

- Place the aesthetic maps that are shared in the `ggplot()` call and place the
  geom-specific aesthetic maps within the `geom_` call
    ```{r, error = TRUE}
    ## Rug plot only on x-axis
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point(mapping = aes(color = drv)) +
      geom_smooth(se = FALSE)
    ```

- **Exercise**: Try adding the OLS line to the scatterplot of petal length vs petal 
  width. Color code by species, but the OLS line should not depend on species.
  
```{r, echo = FALSE, eval = FALSE}
ggplot(data = iris, mapping = aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(mapping = aes(color = Species)) +
  geom_smooth(method = lm, se = FALSE)
```
  
  

## Overplotting

- The variables `cty` and `hwy` are rounded, so result in a lot of overlapping.

- Three ways to counteract this:

    1. If you have small data, use `geom_jitter()`
    2. If you have large data, set the transparency for all points to be low.
    3. If you have a really large dataset, randomly subsample the observational 
       units (chapter 5).
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
      geom_jitter()
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
      geom_point(alpha = 0.1)
    ```

- **Exercise**: Consider the `diamonds` dataset from ggplot2. Load it and then
  plot `carat` vs `price`. Account for overplotting if possible.
  
    ```{r, echo = FALSE, eval = FALSE}
    data("diamonds")
    ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
      geom_point(alpha = 0.01)
    ```
  

# One Quantitative Variable

- A histogram allows us to explore the distribution of a quantitative variable.

- "Distribution" = what values a variable takes and how often it takes those 
  values.

- Bins observations into bins of equal *width*, plots counts on $y$-axis and 
  variable values on $x$-axis.
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy)) +
      geom_histogram()
    ```

- Always play around with the `bins` argument until you get something 
  nice-looking

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy)) +
      geom_histogram(bins = 20)
    ```
    
- You can annotate by a categorical variable by passing the `fill` (best) or
  `color` (worst) aesthetics.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy, fill = drv)) +
      geom_histogram(bins = 20)
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy, color = drv)) +
      geom_histogram(bins = 20)
    ```

- Changing the `position` argument of `geom_histogram()` can help with
  seeing the histograms within all levels of the categorical variable.
  
- You can see all histograms by changing the transparency of the histograms in
  conjunction with `position = "identity"`.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy, fill = drv)) +
      geom_histogram(bins = 20, alpha = 0.3, position = "identity")
    ```

- `position = "dodge"` separates the bins for each category.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy, fill = drv)) +
      geom_histogram(bins = 10, position = "dodge")
    ```

- `position = "fill"` plots proportions of each category within a bin.
    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy, fill = drv)) +
      geom_histogram(bins = 20, position = "fill")
    ```


- **Exercise**: Create a histogram of the price of a diamond, 
  color coding by cut.
  
    ```{r, echo = FALSE, eval = FALSE}
    ggplot(data = diamonds, mapping = aes(x = price, fill = cut)) +
      geom_histogram(bins = 30)
    ```
  

## Density Plot

- A density is a "smoothed" histogram. It can be created with `geom_density()`

- Densities and histograms generally give us identical information.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy)) +
      geom_density()
    ```

- You can plot the densities for the different levels of a categorical variable
  by mapping the `fill` aesthetic
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy)) +
      geom_density(alpha = 0.3)
    ```
    
- You can change the transparency to see all densities.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy, fill = drv)) +
      geom_density(alpha = 0.3)
    ```


# One categorical variable, one quantitative variable.

- The best plot for visualizing the association between a categorical and a
  quantitative variable is the boxplot. Use `geom_boxplot()` for this.
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = drv, y = hwy)) +
      geom_boxplot()
    ```
  
- An alternative is a violin plot:

    ```{r}
    ggplot(data = mpg, mapping = aes(x = drv, y = hwy)) +
      geom_violin()
    ```

- And the beeswarm plot from the package ggbeeswarm.
    ```{r}
    library(ggbeeswarm)
    ggplot(data = mpg, mapping = aes(x = drv, y = hwy)) +
      geom_beeswarm()
    ```

- **Exercise**: Create a boxplot for cut vs log-price. Does anything surprise 
  you? Can you think of another plot that could explain your surprise?

    ```{r, echo = FALSE, eval = FALSE}
    ## Ideal diamonds are lower priced?
    ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
      geom_boxplot() +
      scale_y_log10()
    
    ## But they also tend to be smaller diamonds.
    ggplot(data = diamonds, mapping = aes(x = cut, y = carat)) +
      geom_boxplot() +
      scale_y_log10()
    ```

# Facets

- If you want to create multiple plots from the same data, you can use the 
  `facet_wrap()` or `facet_grid()` arguments.
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point() +
      facet_wrap( ~ class)
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point() +
      facet_grid(. ~ class)
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point() +
      facet_grid(class ~ .)
    ```
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point() +
      facet_grid(class ~ drv)
    ```

- **Exercise**: Create a boxplot of cut vs price for each level of color.

    ```{r, echo = FALSE, eval = FALSE}
    ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
      geom_boxplot() +
      scale_y_log10() +
      facet_wrap( ~ color)
    ```


# Aesthetics for all objects.

- Suppose we like scatterplots with blue points. The correct code to use is:
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
      geom_point(color = "blue")
    ```

- The `"blue"` must be outside of the `aes()` call. This tells ggplot2 to set
  the aesthetic manually, instead of via a map from a variable.
  
- If you put `"blue"` inside of the `aes()` call, then ggplot2 will assume that
  you are creating a categorical variable where every observational unit has
  the value `"blue"`.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = "blue")) +
      geom_point()
    ```

- I often change the aesthetics for histograms.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy)) +
      geom_histogram(fill = "white", color = "black")
    ```
    
- I also often use the `alpha` argument frequently (see overplotting above).

# Themes

- The gray background is nice for presentations, but bad for publications. 
  ggplot2 has a lot of themes that you can use. I think `theme_bw()` is the 
  best.
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_bw()
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_classic()
    ```
    
- Even more themes are available in the ggthemes package.

    ```{r}
    library(ggthemes)
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_excel()
    ```
    
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_fivethirtyeight()
    ```

- **Exercise**: Create a boxplot of cut vs price for each level of color. Make
  the boxplots all orange and use the black and white theme.
  
    ```{r, echo = FALSE, eval = FALSE}
    ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
      geom_boxplot(fill = "orange") +
      scale_y_log10() +
      facet_wrap( ~ color) +
      theme_bw()
    ```
  
  

# Scales

- "Scales control how a plot maps data values to the visual
  values of an aesthetic. To change the mapping, add a
  custom scale."
  
- E.g.: The the scale of colors determines what color is the "largest" and 
  which is the "smallest."
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = cty)) +
      geom_point() +
      scale_color_continuous(low = "black", high = "white")
    ```
  
- E.g. The scale of an axis (log vs normal)

    ```{r}
    ggplot(data = diamonds, mapping = aes(x = price, y = carat)) +
      scale_y_log10() +
      scale_x_log10() +
      geom_point(alpha = 0.01)
    ```

# Colorblind  safe pallettes

- When you publish, it's polite to use colorblind safe palettes. You can do this
  easily with the ggthemes package.
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_bw() +
      scale_color_colorblind()
    ```

- Use `scale_color_colorblind()` to change the scale of the `color` aesthetic.
  Use `scale_fill_colorblind()` to change the scale of the `fill` aesthetic,
  etc...

    ```{r}
    ggplot(data = mpg, mapping = aes(x = hwy, fill = drv)) +
      geom_histogram(position = "dodge", bins = 15) +
      theme_bw() +
      scale_fill_colorblind()
    ```
    
# Labels

- You can set the axis labels manually with `xlab()` and `ylab()`.

- Titles may be added with `ggtitle()`.

- Legend titles need to be modified with a scale argument.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_bw() +
      scale_color_colorblind(name = "Car Class") +
      xlab("Displacement") +
      ylab("Highway") +
      ggtitle("Displacement vs Highway")
    ```
    
- If you don't want colorblind scale, you can use `scale_color_discrete()` or
  `scale_fill_discrete()` or `scale_color_continuous()` or `scale_fill_continuous()`
  etc...
  
    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_bw() +
      scale_color_discrete(name = "Car Class") +
      xlab("Displacement") +
      ylab("Highway") +
      ggtitle("Displacement vs Highway")
    ```
  

    
# Saving Plots

- To save a plot, you can either print it and then use the RStudio functionality.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
      geom_point() +
      theme_bw() +
      scale_color_colorblind() ->
      pl
    ```

- It's better to automatic this process using the `ggsave()` function
    ```{r}
    ggsave(filename = "my_first_plot.pdf",
           plot = pl,
           width = 6, 
           height = 4,
           family = "Times")
    ```


- You can also do this manually using `pdf()`:
    ```{r}
    pdf(file = "my_second_plot.pdf", width = 6, height = 4, family = "Times")
    print(pl)
    dev.off()
    ```
    
- `pdf()` opens up a connection to a new pdf file. `print(pl)` prints our saved
  `gg` object to that pdf file (instead of the R graphics device). 
  `dev.off()` closes the connection.
  
- **Exercise**: Create a boxplot of cut vs price for each level of color. Make
  the boxplots all orange and use the black and white theme. Save this plot to
  the output folder using `ggsave()`.
  
    ```{r, echo = FALSE, eval = FALSE}
    ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
      geom_boxplot(fill = "orange") +
      scale_y_log10() +
      facet_wrap( ~ color) +
      theme_bw() ->
      pl
    ggsave(filename = "./box_cut_price_color.pdf", plot = pl)
    ```

    

