---
title: 'Lab 03: ggplot2'
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
params:
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = params$solutions, 
                      eval       = params$solutions,
                      fig.align  = "center",
                      fig.height = 3, 
                      fig.width  = 5)
```

# Chick Weight

The `ChickWeight` data frame contains information on the effect of diet
on the weight of chicks in an experiment from 1990. Load it into R and
read about it with:
```{r, echo = TRUE, eval = FALSE}
data("ChickWeight")
help("ChickWeight")
```

Reproduce the following plots exploring the effect of diet on weight.
```{r, eval = TRUE, message = FALSE}
library(tidyverse)
library(ggthemes)
ggplot(ChickWeight, aes(x = Time, y = weight, color = Diet, group = Chick)) +
  geom_line(alpha = 1/2) +
  scale_color_colorblind() +
  xlab("Time (days)") +
  ylab("Weight (grams)") +
  theme_bw()
```

```{r, eval = TRUE}
ggplot(ChickWeight, aes(x = factor(Time), y = weight, color = Diet)) +
  geom_boxplot(outlier.size = 0.3) +
  xlab("Time (days)") +
  ylab("Weight (grams)") +
  theme_bw()
```

# Insect Sprays

Read about the `InsectSprays` data frame by typing into R:
```{r, echo = TRUE, eval = FALSE}
data("InsectSprays")
help("InsectSprays")
```

These data contain one categorical and one quantitative variable. We can
explore their association through a boxplot. Reproduce the plot below:
```{r, eval = TRUE}
ggplot(InsectSprays, aes(x = spray, y = count)) +
  geom_boxplot(fill = "red", alpha = 1/2) +
  xlab("Spray Type") +
  ylab("Insect Count") +
  theme_bw()
```

Since there are few observations, you decide to just plot the points, but you
want to account for overlapping points. Try
to approximate this plot (you won't be able to get it *exactly* the same):
```{r, eval = TRUE}
set.seed(1)
ggplot(InsectSprays, aes(x = spray, y = count)) +
  xlab("Spray Type") +
  ylab("Insect Count") +
  geom_jitter(width = 0.1, height = 0) +
  theme_bw()
```


# Michelson Speed of Light

Load in and read about the `morley` data frame in R:
```{r, eval = FALSE, echo = TRUE}
data("morley")
help("morley")
```

Do the different experiments predict different speeds of light? Reproduce the
following plot:
```{r, eval = TRUE}
ggplot(morley, aes(x = factor(Expt), y = Speed)) +
  geom_boxplot() +
  xlab("Experiment Number") +
  ylab("Speed") +
  theme_bw()
```

Is there a trend in the experimental runs? Reproduce this plot in R:
```{r, fig.height=9, eval = TRUE}
## no trend observed
ggplot(morley, aes(x = Run, y = Speed)) +
  facet_grid(Expt ~ .) +
  geom_line() +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white")) +
  scale_x_continuous(breaks = 1:20)
```


# Rock

Read about the `rock` data frame by:
```{r, eval = FALSE, echo = TRUE}
data("rock")
help("rock")
```

Does area seem to be associated with permeability? Justify with a plot.
```{r}
## hard to say
ggplot(rock, aes(x = area, y = perm)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```




