---
title: "Lattice R Graphics Cheat Sheet"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 3, fig.align = "center")
```


# Abstract:
I reproduce some of the plots from Rstudio's [ggplot2](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) cheat sheet using just the lattice R package.

```{r, echo=FALSE}
set.seed(1)
```

Before we begin, load the lattice package in R:
```{r}
library(lattice)
```

We'll use the `mpg` dataset from ggplot2:
```{r}
library(ggplot2)
data("mpg")
```

# General Considerations

- Look at the help page of `xyplot()`, which contains a lot of details on many of
the high-level plotting functions in lattice.
- Most of the high-level plotting functions use a formula as the first input with the 
  variable on the y-axis to the left of the variable on the x-axis, separated
  by a tilde: `y ~ x`.
- If you want to layer multiple geometric objects (scatterplots, loess smoother, 
rugplot, etc) onto one plot, you need to use the `panel` argument, where you
specify the plotting functions such as `panel.xyplot()` or `panel.loess()` or
`panel.rug()`, etc. Examples of this implementation are below.

# One Variable

Density Plot
```{r}
densityplot( ~ hwy, data = mpg)
```

Histogram
```{r}
histogram( ~ hwy, data = mpg)
```


# Discrete

Barplot
```{r}
barchart(mpg$drv, horizontal = FALSE)
```

\vspace{20mm}

## Two Variables

## Continuous $X$, Continuous $Y$

Scatterplot
```{r}
xyplot(hwy ~ cty, data = mpg)
```

Add a rug plot

```{r}
xyplot(hwy ~ cty, data = mpg,
       panel = function(x, y) {
         panel.xyplot(x, y)
         panel.rug(x, y)
       })
```

\vspace{20mm}

Add a loess smoother plot

```{r}
xyplot(hwy ~ cty, data = mpg,
       panel = function(x, y) {
         panel.xyplot(x, y)
         panel.rug(x, y)
         panel.loess(x, y)
       })
```



Add some text

```{r}
xyplot(hwy ~ cty, data = mpg,
       panel = function(x, y) {
         panel.xyplot(x, y)
         panel.rug(x, y)
         panel.loess(x, y)
         panel.text(x = 15, y = 40, label = "some text")
       })
```


## Discrete X, Continuous Y

Boxplot

```{r}
bwplot(hwy ~ class, data = mpg)
```

## Continuous Function

Line Plot

```{r}
x <- seq(-2, 2, length = 100)
y <- x ^ 2
xyplot(y ~ x, data.frame(x, y), type = "l")
```

\vspace{20mm}

## Color Coding and Legend Title

```{r}
xyplot(cty ~ hwy, data = mpg, groups = drv, auto.key = list(title = "Drive"))
```

## Faceting

Use the `|` command in the formula of `xyplot()` or `bwplot()` or `histogram()` 
or `barchart()` to facet.

```{r, fig.width=6, fig.height=4}
xyplot(cty ~ hwy | drv, data = mpg)
```

```{r, fig.height=8, fig.width=6}
xyplot(cty ~ hwy | drv + fl, data = mpg)
```
