---
title: "`qplot` R Graphics Cheat Sheet"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 3, fig.height = 2, fig.align = "center")
```

# Abstract:
I reproduce some of the plots from Rstudio's [ggplot2](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) cheat sheet using just the `qplot` function.

```{r, echo=FALSE}
set.seed(1)
```

Before using `qplot` in a new R session, always first load the ggplot2 library.
```{r}
library(ggplot2)
```

I use this dataset
```{r}
data(mpg, package = "ggplot2")
```

# General Considerations

The main options that I use are

- Options for `"geom"` argument:
    - `"point"`: Makes scatterplots.
    - `"line"`: Makes a line plot.
    - `"histogram"`: Makes a histogram.
    - `"boxplot"`: Makes a boxplot.
    - `"density"`: Makes the density plot.
    - `"bar"`: First tabulates frequencies of each value, then makes a barplot.
    - `"smooth"`: Fits a smooth line to a cloud of points and plots the output.
    - `"dotplot"`: Makes a dotplot.

`qplot` has other arguments that control the way the plot looks. You should read about these arguments. In particular, read carefully the help page `?qplot`. Useful ones are:

- `data`: Specify the dataframe that all variables belong to.
- `main`: This controls the title.
- `xlab`, `ylab`: These control the x and y axis labels.
- `color`: Controls the color of the lines/points.
- `fill`: Controls the color of areas (e.g. for histograms).
- `size`: Controls the size of points.
- `shape`: The shape of points (`"circle"`, `"square"`, `"triangle"`, etc...)
- `alpha`: Controls the level of transparency of points/lines/fills.
- `lwd`: Line width.
- `lty`: Line type (`"solid"`, `"dashed"`, `"dotted"`, etc...).
- `facets`: Split up the data into multiple plots.

If you want to make all points the same shape/size/color, you need to enclose the size/shape/color using the function `I()`.

If a variable is being treated as continuous rather than categorical, you need to enclose that variable in a `factor()` function call.

# One Variable

## Continuous

Density plot
```{r}
qplot(x = mpg$hwy, geom = "density")
```

Histogram
```{r}
qplot(mpg$hwy, geom = "histogram", bins = 10)
```

Make the bin lines black and the fill white.

```{r}
qplot(mpg$hwy, geom = "histogram", bins = 10, color = I("black"), fill = I("white"))
```


## Discrete
Barplot
```{r}
qplot(mpg$drv, geom = "bar")
```

# Two Variables

## Continuous X, Continuous Y

Scatterplot
```{r}
qplot(mpg$cty, mpg$hwy, geom = "point")
```

\newpage
Jitter points to account for overlaying points.
```{r}
x <- jitter(mpg$cty)
y <- jitter(mpg$hwy)
qplot(x, y, geom = "point")
```

Add a rug plot
```{r}
qplot(x, y, geom = "point") + 
  geom_rug()
```

Add a Loess Smoother
```{r}
qplot(x, y, geom = "point") + 
  geom_rug() + 
  geom_smooth()
```

Add text to a plot
```{r}
qplot(x, y, geom = "point") + 
  geom_rug() + 
  geom_smooth() + 
  annotate(geom = "text", x = 15, y = 40, label = "some text")
```


## Discrete X, Continuous Y

Boxplot
```{r}
qplot(x = mpg$class, y = mpg$hwy, geom = "boxplot")
```

\newpage

## Continuous Function

Line plot
```{r}
x <- seq(-2, 2, length = 100)
y <- x^2
qplot(x, y, geom = "line")
```

# Color Coding and Legends
Color code a scatterplot by a categorical variable and add a legend.
```{r}
x <- jitter(mpg$hwy)
y <- jitter(mpg$cty)
z <- factor(mpg$drv)
qplot(x, y, color = z)
```

\newpage

# Changing a legend title

```{r}
qplot(x, y, color = z) + scale_color_discrete(name = "New Name1")
qplot(x, y, shape = z) + scale_shape_discrete(name = "New Name2")
```

# The `data` argument
If all variables you are using in `qplot()` belong to the same dataframe, then you can specify the dataframe as the "data" argument and you don't need to use the "`$`" symbol.

```{r}
qplot(cty, hwy, color = drv, data = mpg, geom = "point")
```

# Faceting

You can facet by a categorical variable using the `facets` argument.

The variable to the left of the tilde ("`~`") indexes the row facets, the variable to the right of the tilde indexes the column facets. Using a dot ("`.`") in place of a variable means that there will only be one row/column facet.

```{r}
qplot(cty, hwy, data = mpg, facets = . ~ drv, geom = "point")
qplot(cty, hwy, data = mpg, facets = drv ~ ., geom = "point")
```
```{r, fig.height=2.7, fig.width=3.5}
qplot(cty, hwy, data = mpg, facets = fl ~ drv, geom = "point")
```





