---
title: "Parsers"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Change character vectors into other types using parsers.
- Parsers and reader.
- Chapter 11 of [RDS](https://r4ds.had.co.nz/)

# Motivation

- Suppose you have the following data frame
    ```{r}
    suppressPackageStartupMessages(library(tidyverse))
    dfdat <- tribble(
      ~date,        ~time,      ~number, ~factor, ~logical,
      ##----------  ----------  -------  -------  --------
      "12-01-1988", "10:10:02", "2",     "A",     "TRUE",
      "11-12-1987", "11:10:57", "4",     "A",     "TRUE",
      "02-03-1989", "10:10:25", "6",     "B",     "FALSE",
      "06-03-1982", "22:10:55", "2",     "B",     "TRUE",
      "09-21-1981", "10:10:02", "1",     "A",     "FALSE"
      )
    dfdat
    ```

- How do we convert the characters to the types we want? Parse!

# Parsing dates and times

- See [`{lubridate}`](https://data-science-master.github.io/lectures/02_dates/02_dates.html) notes.
    
# Parsing Numbers

- `parse_double()` and `parse_integer()` expect strict numbers and will fail
   if there is anything non-number-like.
  
    ```{r}
    parse_double("2.11")
    parse_double("$2.11")
    
    parse_integer("2")
    parse_integer("2%")
    ```
    
- `parse_number()` removes non-numeric characters.

    ```{r}
    parse_number("$2.11")
    parse_number("2%")
    ```
    
- You can change the grouping variable from "," to "." with

    ```{r}
    parse_number("2.555,11", 
                 locale = locale(grouping_mark = ".", 
                                 decimal_mark = ","))
    ```
    
- Example:

    ```{r}
    dfdat %>%
      mutate(number = parse_number(number))
    ```


# Parsing other types

- `parse_logical()` and `parse_factor()` and `parse_string()` are pretty 
  self-explanatory.
  
  
    ```{r}
    dfdat %>%
      mutate(factor = parse_factor(factor))
    ```
    
    ```{r}
    dfdat %>%
      mutate(logical = parse_logical(logical))
    ```

# Parsing and readr

- When you specify `col_types` in `read_csv()`, those are wrappers for parsers.

    ```{r}
    read_csv("../../data/estate.csv",
             col_types = cols(
               Price   = col_double(),
               Area    = col_double(),
               Bed     = col_double(),
               Bath    = col_double(),
               AC      = col_logical(),
               Garage  = col_double(),
               Pool    = col_logical(),
               Year    = col_double(),
               Quality = col_factor(),
               Style   = col_factor(),
               Lot     = col_double(),
               Highway = col_logical()
               )) ->
      estate
    estate
    ```


    



