---
title: "tidyr Lab"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
params: 
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$solutions, eval = params$solutions)
```

# Exercise 1

Tidy the data frame `ex0724` from the `Sleuth3` package. You can read about this data frame by typing `help(ex0724)` after loading Sleuth3.

```{r}
library(Sleuth3)
data("ex0724")
ex0724 %>%
  gather(Denmark:USA, key = "country", value = "proportion", na.rm = TRUE)
```

# Exercise 2

Load in and tidy the `pred` data frame: https://dcgerard.github.io/stat_412_612/data/preg.csv

Make the coding for any new variables nice.

Save the tidied data frame in the output folder.

```{r}
suppressPackageStartupMessages(library(tidyverse))
preg <- read_csv("https://dcgerard.github.io/stat_412_612/data/preg.csv")

preg %>%
  gather(treatmenta, treatmentb, key = "Treatment", value = "Value") %>%
  mutate(Treatment = recode(Treatment, "treatmenta" = "A", "treatmentb" = "B")) ->
  preg2
preg2
```

# Exercise 3

Load in and tidy the `pew` data frame: https://dcgerard.github.io/stat_412_612/data/pew.csv

Values in the cells are counts of people who adhere to a particular religion and contain a particular salary range.

Save the tidied data in the output folder.

```{r}
pew <- read_csv("https://dcgerard.github.io/stat_412_612/data/pew.csv")
pew %>%
  gather(-religion, key = "Salary", value = "Count") ->
  pew2
pew2
```


# Exercise 4

Load in and tidy the `tb` data frame: https://dcgerard.github.io/stat_412_612/data/tb.csv

The column names specify both the sex (`m` = male, `f` = femail) and age range (`04` = 0 to 4, `514` = 5 to 14, `014` = 0 to 14, `1524` = 15 to 24, `2534` = 25 to 34, `3544` = 35 to 44, `4554` = 45 to 54, `4464` = 55 to 64, `65` = >=65, `u` = unknown). The values in the cells are counts.

Save the tidied data in the output folder.

```{r}
tb <- read_csv("https://dcgerard.github.io/stat_412_612/data/tb.csv")

tb %>%
  gather(-iso2, -year, key = "sex_age", value = "counts", na.rm = TRUE) %>%
  separate(col = sex_age, into = c("sex", "age"), sep = 1) ->
  tb2
tb2
```

# Exercise 5 

Load in and tidy the `weather` data frame: https://dcgerard.github.io/stat_412_612/data/weather.csv

- A column header that begins with a "d" is a day. 
- For `element`, `tmax` = maximum temperature, `tmin` = minimum temperature. 
- The values in the cells are the maximum or minimum temperatures for given days of the year.

Save the tidied data in the output folder.

```{r}
weather <- read_csv("https://dcgerard.github.io/stat_412_612/data/weather.csv")
weather %>%
  gather(-id, -year, -month, -element, key = "day", value = "temperature", na.rm = TRUE) %>%
  spread(key = element, value = "temperature") %>%
  mutate(day = parse_number(str_replace(day, "d", ""))) %>%
  arrange(year, month, day) ->
  weather2
weather2
```

# Exercise 6


Load in and tidy the `wine` data frame: https://dcgerard.github.io/stat_412_612/data/wine.csv

```{r}
wine <- read_csv2("https://dcgerard.github.io/stat_412_612/data/wine.csv")
```



