---
title: "dplyr Lab"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
params: 
  solutions: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$solutions, eval = params$solutions)
```

- The following exercises are all from Section 5.7.1 of [RDS](https://r4ds.had.co.nz/).

- Use the `flights` data frame from the nycflights13 package.

1. Which plane (`tailnum)` has the worst departure delay record?

    ```{r}
    library(tidyverse)
    library(nycflights13)
    data("flights")
    ```

    ```{r}
    flights %>%
      group_by(tailnum) %>%
      summarize(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
      arrange(mean_dep_delay) %>%
      slice(1)
    ```

2. What time of day should you fly if you want to avoid delays as much as 
  possible?
  
    ```{r}
    flights %>%
      group_by(hour) %>%
      summarize(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
      ggplot(mapping = aes(x = hour, y = mean_dep_delay)) +
      geom_line() +
      xlab("Hour") +
      ylab("Departure Delay (min)") +
      theme_bw()
    ```

    ```{block}
    You should leave at 5 AM.
    ```

3. For each destination, compute the total minutes of arrival delay. For each 
  flight, compute the proportion of the arrival delay for its destination.
  
    ```{r}
    flights %>%
      group_by(dest) %>%
      mutate(dest_delay = sum(arr_delay, na.rm = TRUE),
             prop_delay = arr_delay / dest_delay) %>%
      ungroup() %>%
      select(dest, dest_delay, prop_delay)
    ```
  

4. Delays are typically temporally correlated: even once the problem that 
   caused the initial delay has been resolved, later flights are delayed to 
   allow earlier flights to leave. Using `lag()`, explore how the departure delay 
   of a flight is related to the delay of the immediately preceding flight.
   
    ```{r}
    flights %>%
      group_by(origin) %>%
      arrange(origin, year, month, day, hour, minute) %>%
      mutate(next_dep_del = lag(dep_delay)) %>%
      ungroup() ->
      sumdat
    
    ## Moderate/weak correlation
    sumdat %>%
      group_by(origin) %>%
      summarize(cor = cor(dep_delay, next_dep_del, use = "pairwise.complete.obs"))
    
    
    sumdat %>%
      sample_n(size = 5000) %>%
      ggplot(mapping = aes(x = dep_delay, y = next_dep_del)) +
      geom_point(alpha = 0.3) +
      theme_bw() +
      xlab("Departure Delay") +
      ylab("Next Departure Delay") +
      geom_smooth(se = FALSE)
    ```
   

5. Look at each destination. Can you find flights that are suspiciously fast?
  (i.e. flights that represent a potential data entry error).
  Compute the air time of a flight relative to the shortest flight to that 
  destination. Which flights were most delayed in the air?
  
    ```{r}
    flights %>%
      group_by(dest) %>%
      mutate(ref_air_time = air_time - min(air_time, na.rm = TRUE),
             mean_air_time = mean(air_time, na.rm = TRUE),
             sd_air_time = sd(air_time, na.rm = TRUE), 
             z_air_time = (air_time - mean_air_time) / sd_air_time) %>%
      select(dest, air_time, ref_air_time, z_air_time, mean_air_time, sd_air_time) ->
      df_air
    
      ggplot(data = df_air, mapping = aes(x = z_air_time)) +
        geom_histogram()
      
    df_air %>%
      arrange(z_air_time)
    
    df_air %>%
      mutate(rescale_air_time = air_time / min(air_time, na.rn = TRUE)) %>%
      ungroup() %>%
      arrange(desc(rescale_air_time))
    ```
  

6. Find all destinations that are flown by at least two carriers. (hint: use `n_distinct()`)
  
    ```{r}
    flights %>%
      group_by(dest) %>%
      summarize(ncarr = n_distinct(carrier)) %>%
      filter(ncarr >= 2)
    ```
