---
title: "Strings and Regular Expressions"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Manipulating strings with the stringr package.
- Regular expressions
- Chapter 14 of [RDS](https://r4ds.had.co.nz/).
- [Work with Strings Cheatsheet](https://www.rstudio.com/resources/cheatsheets/#stringr).

# Strings

- In R, strings (also called "characters") are created and displayed within quotes:

    ```{r}
    x <- "I am a string!"
    x
    ```

- Anything within quotes is a string, even numbers!
    ```{r, error=TRUE}
    y <- "3"
    class(y)
    ```

- You can have a vector of strings.

    ```{r}
    x <- c("I", "am", "a", "string", "vector")
    x[2:3]
    ```
    
- The backslash `"\"` means that what is after the backslash is special is some 
  way. For example, if you want to put a quotation mark in a string, you can "escape" the
  quotation mark with a backslash.
  
    ```{r}
    x <- "As Tolkein said, \"Not all those who wonder are lost\""
    writeLines(x)
    ```
    
- Above, `writeLines()` will print out the string itself. `print()` will print 
  out the printed representation of the string (with backslashes and all).
  
    ```{r}
    print(x)
    ```
  
  
- `"\n"` represents a new line.

    ```{r}
    x <- "Not all those\nwho wonder are lost."
    writeLines(x)
    ```
    
- `"\t"` represents a tab.

    ```{r}
    x <- "Not all those\twho wonder are lost."
    writeLines(x)
    ```
    
- You can add any unicode character with a `\u` followed by the hexidecimal 
  [unicode representation](https://en.wikipedia.org/wiki/List_of_Unicode_characters) 
  of that character.
  
    ```{r}
    mu <- "\u00b5"
    writeLines(mu)
    ```
  
# stringr

- The stringr package contains a lot of convenience functions for manipulating
  strings (and they are a lot more user friendly than base R's string 
  manipulation functions like `grep()` and `gsub()`).

- stingr is not part of the tidyverse so you have to load it separately.

    ```{r, message = FALSE}
    library(tidyverse)
    library(stringr)
    ```
    

  
  

