---
title: "Strings and Regular Expressions"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
```

# Learning Objectives

- Manipulating strings with the stringr package.
- Regular expressions
- Chapter 14 of [RDS](https://r4ds.had.co.nz/).
- [Work with Strings Cheatsheet](https://www.rstudio.com/resources/cheatsheets/#stringr).

# Strings

- In R, strings (also called "characters") are created and displayed within quotes:

    ```{r}
    x <- "I am a string!"
    x
    ```

- Anything within quotes is a string, even numbers!
    ```{r, error=TRUE}
    y <- "3"
    class(y)
    ```

- You can have a vector of strings.

    ```{r}
    x <- c("I", "am", "a", "string", "vector")
    x[2:3]
    ```
    
- The backslash `"\"` means that what is after the backslash is special in some 
  way. For example, if you want to put a quotation mark in a string, you can "escape" the
  quotation mark with a backslash.
  
    ```{r}
    x <- "As Tolkein said, \"Not all those who wonder are lost\""
    writeLines(x)
    ```
    
- Above, `writeLines()` will print out the string itself. `print()` will print 
  out the printed representation of the string (with backslashes and all).
  
    ```{r}
    print(x)
    ```
  
  
- `"\n"` represents a new line.

    ```{r}
    x <- "Not all those\nwho wonder are lost."
    writeLines(x)
    ```
    
- `"\t"` represents a tab.

    ```{r}
    x <- "Not all those\twho wonder are lost."
    writeLines(x)
    ```
    
- You can add any Unicode character with a `\u` followed by the hexadecimal 
  [unicode representation](https://en.wikipedia.org/wiki/List_of_Unicode_characters) 
  of that character.
  
    ```{r}
    mu <- "\u00b5"
    writeLines(mu)
    ```
  
# stringr Intro

- The stringr package contains a lot of convenience functions for manipulating
  strings (and they are a lot more user friendly than base R's string 
  manipulation functions like `grep()` and `gsub()`).

- stingr is not part of the tidyverse so you have to load it separately.

    ```{r, message = FALSE}
    library(tidyverse)
    library(stringr)
    ```
    
- All of stringr's functions begin with "`str_`", so you can press tab after
  typing "`str_`" and a list of possible string manipulation functions will
  pop up (in RStudio).
  
- For example, to get the number of characters in a string, use `str_length()`.

    ```{r}
    str_length("Upon the hearth the fire is red,")
    ```

    
## Combining Strings

- Combine strings with `str_c()`.
    ```{r}
    x <- "Faithless is he that says"
    y <- "farewell when the road darkens."
    str_c(x, y)
    ```

- The default is to separate strings by nothing, but you can use `sep` to change
  the separator.
    ```{r}
    str_c(x, y, sep = " ")
    ```
    
- Just like `c()`, `str_c()` can take multiple arguments.

    ```{r}
    str_c("Short", "cuts", "make", "long", "delays.", sep = " ")
    ```
    
- If you provide `str_c()` a vector of arguments, it will vectorize the 
  combining unless you provide a `collapse` argument.
  
    ```{r}
    x <- c("Short", "cuts", "make", "long", "delays.")
    str_c(x, "LOTR", sep = " ")
    str_c(x, collapse = " ")
    ```

- Combining with `NA` results in `NA`:
    ```{r}
    str_c("Faithless is he that says", NA)
    ```
    

## Extracting substrings

- `str_sub()` extracts a substring between the location of two characters.
    ```{r}
    x <- "The Road goes ever on and on"
    str_sub(x, start = 2, end = 6)
    ```
    
- Replace substrings with assignment
    ```{r}
    str_sub(x, start = 2, end = 6) <- " Tolkein "
    x
    ```
    
- **Exercise**: Reproduce this quote

    > But under a tall tree I will lie,
      And let the clouds go sailing by.

    with these strings
    ```{r}
    w <- "But under a tall tree"
    x <- "FRELL I will lie"
    y <- "and let clouds go"
    z <- "sailing by."
    ```
    
    ```{r, echo = FALSE, eval = FALSE}
    str_c(w, str_sub(x, 7, str_length(x)), y, z, sep = " ")
    ```
    

# Regular Expressions

- Regular expressions (regex or regexp) is a syntax for pattern matching in strings.

- We'll use `str_replace()` and `str_replace_all()` to demonstrate using regex 
  in stringr. These functions search for a pattern and then replace it with
  another string.
  
- But wherever there is a `pattern` argument in a stringr function, you can 
  use regex (to extract strings, get a logical if there is a match, etc...).
  
- Basic usage: finds exact match of string.

    ```{r}
    x <- "Ho! Ho! Ho! to the bottle I go to heal my heart and drown my woe."
    str_replace_all(x, "hea", "XX")
    ```

- A period "`.`" matches any character.

    ```{r}
    str_replace_all(x, "hea.", "XX")
    ```
    
- You can "escape" a period with two backslashes "`\\`" to match periods.

    ```{r}
    str_replace_all(x, ".", "X") ## Matches everything
    str_replace_all(x, "\\.", "X") ## Matches the only period
    ```

- To match a backslash, you need four backslashes (to escape the escape).

    ```{r}
    y <- "Rain\\may\\fall\\and\\wind\\may\\blow"
    writeLines(y)
    str_replace_all(y, "\\\\", "XX")
    ```


- *Important note*: The actual regular expressions above are strings themselves,
  and so you view them with `writeLines()`. So using "`\\.`" as the pattern
  argument in R results in the regular expression "`\.`".

- **Exercise**: Construct a regular expression to match this string:
    ```{r, echo = FALSE}
    writeLines("\\.\\.\\.")
    ```

- **Exercise**: Use one function call to replace `"back"` and `"lack"` with `"foo"`.

    ```{r}
    x <- "but better is Beer if drink we lack, and Water Hot poured down the back."
    ```

    ```{r, echo = FALSE, eval = FALSE}
    str_replace_all(x, ".ack", "foo")
    ```
    
## Anchoring

- You can anchor the pattern to only match the start or end of a string.

    - `^` matches only the start of a string.
    - `$` matches only the end of a string.

    ```{r}
    x <- c("But", "under", "a", "tall", "tree", "I", "will", "lie")
    str_replace(x, "^t", "XX")
    str_replace(x, "t$", "XX")
    ```
    
- Use both to match only a complete string.

    ```{r}
    x <- c("apple pie", "apple", "apple cake")
    str_replace_all(x, "apple", "XX")
    str_replace_all(x, "^apple$", "XX")
    ```
    
- **Exercise**: Use `str_replace()` to replace all four letter words beginning
  with an `"a"` with `"foo"` in the following list
    ```{r}
    x <- c("apple", "barn", "ape", "cart", "alas", "pain", "ally")
    ```
    
    ```{r, echo = FALSE, eval = FALSE}
    str_replace(x, "^a...$", "foo")
    ```
    
## Special Characters

- We'll use this character vector for practice:
    ```{r}
    x <- c("Abba: 555-1234", "Anna: 555-0987", "Andy: 555-7654")
    ```

- `\\d`: matches any digit.
    ```{r}
    str_replace(x, "\\d\\d\\d-\\d\\d\\d\\d", "XXX-XXXX")
    ```
- `\\s`: matches any white space (e.g. space, tab, newline).
    ```{r}
    str_replace(x, "\\s", "X")
    ```
- `[abc]`: matches `a`, `b`, or `c`.
    ```{r}
    str_replace(x, "A[bn][bn]a", "XXXX")
    ```
- `[^abc]`: matches anything except `a`, `b`, or `c`.
    ```{r}
    str_replace(x, "A[^b]", "XXXX")
    ```
- `abc|xyz`: matches either `abc` or `xyz`. This is called *alternation*
- You can use parentheses to control where the alternation occurs.
    - `a(bc|xy)z` matches either `abcz` or `axyz`.
    ```{r}
    str_replace(x, "An(na|dy)", "XXXX")
    ```

- To ignore case, place a `(?i)` before the regex.
    ```{r}
    str_replace("AB", "ab", "X")
    str_replace("AB", "(?i)ab", "X")
    ```

- **Exercise**: Create separate regular expressions to find all words that:

    1. Start with a vowel. Test on 
    ```{r}
    x1 <- c("abba", "cat", "eal", "ion", "oops", "Uganda", "Anna", "dog")
    ```

    2. That end in consonants. (Hint: thinking about matching "not"-vowels.) 
       test on
    ```{r}
    x2 <- c("bob", "Anna", "dog")
    ```

    3. End with `ed`, but not with `eed`. Test on
    ```{r}
    x3 <- c("tired", "need", "bad", "rod")
    ```
    

    4. End with `ing` or `ise`. Test on
    ```{r}
    x4 <- c("paradise", "firing", "jaded", "kin")
    ```
    
    
    ```{r, echo = FALSE, eval = FALSE}
    #1
    str_replace_all(x1, "^[aeiouAEIOU]", "XX")
    #2
    str_replace_all(x2, "[^aeiouAEIOU]$", "XX")
    #3
    str_replace(x3, "[^e]ed$", "XX")
    #4
    str_replace(x4, "(ing|ise)$", "XX")
    ```
    
## Repetition

- Can match a pattern multiple times in a row:

    - `?`: 0 or 1
    - `+`: 1 or more
    - `*`: 0 or more
    
    ```{r}
    x <- c("A", "AA", "AAA", "AAAA", "B", "BB")
    str_replace_all(x, "^A?", "X")
    str_replace_all(x, "^A+", "X")
    str_replace_all(x, "^A*", "X")
    ```
    
- A more realistic example:
    ```{r}
    str_replace_all("color and colour", "colou?r", "X")
    ```

- Control exactly how many repetitions allowed in a match:

    - `{n}`: exactly `n`.
    - `{n,}`: `n` or more.
    - `{0,m}`: at most `m`.
    - `{n,m}`: between `n` and `m`.
  
    ```{r}
    str_replace_all(x, "A{2}", "X")
    str_replace_all(x, "A{2,}", "X")
    str_replace_all(x, "A{0,2}", "X")
    str_replace_all(x, "A{3,4}", "X")
    ```
    
- Regex will automatically match the longest string possible.

    ```{r}
    str_replace("AAAA", "A*", "X")
    ```
    
- **Exercise**: Create regular expressions to find all words that:

    1. Start with three consonants. Test on
    ```{r}
    x1 <- c("string", "priority", "value", "distinction")
    ```
    
    2. Have three or more vowels in a row. Test on
    ```{r}
    x2 <- c("honorific", "delicious", "priority", "queueing")
    ```
    
    3. Have two or more vowel-consonant pairs in a row. Test on
    ```{r}
    x3 <- c("honorific", "sam", "prior")
    ```
    
    ```{r, eval = FALSE, echo = FALSE}
    str_replace_all(x1, "^[^aeiouAEIOU]{3}", "X")
    str_replace_all(x2, "[aeiouAEIOU]{3,}", "X")
    str_replace_all(x3, "([aeiouAEIOU][^aeiouAEIOU]){2,}", "X")
    ```
    
## Grouping and Backreferences
  
- Parentheses create a numbered group that you can then back reference with
  `\\1` for the match in the first parentheses, `\\2` in the second 
  parentheses, etc...

    ```{r}
    str_replace("cococola", "(..)\\1", "pepsi")
    str_replace("banana", "([aeiou][^aeiou])\\1", "XX")
    ```

# stringr tools

- There are a lot of tools, so we'll go over them briefly and do an exercise
  where you can use them in more detail.
  
- `str_to_lower()` and `str_to_upper()` convert all letters to lower or capital
  case.
    ```{r}
    x <- "Deeds will not be less valiant because they are unpraised."
    str_to_lower(x)
    str_to_upper(x)
    ```

- `str_detect()`: Returns `TRUE` if a regex pattern matches a string and `FALSE`
  if it does not. Very useful for filters.
  
    ```{r}
    ## Get all John's and Joe's from the Lahman dataset
    library(Lahman)
    data("Master")
    Master %>%
      filter(str_detect(nameFirst, "^Jo(e|hn)$")) %>%
      select(nameFirst) %>%
      head()
    ```
  
- `str_subset()`: Returns the words where there is a match. Not often as useful
  as `str_detect()` because you don't use it in data frames that often.
  
    ```{r}
    str_subset(Master$nameFirst, "^Jo(e|hn)$") %>%
      head()
    ```
  
- `str_count()`: Counts the occurrence of a match within a string.

    ```{r}
    str_count(c("banana", "coco"), "[^aeiou][aeiou]")
    ```
    
    They count *non-overlapping* matches
    
    ```{r}
    str_count("abababa", "aba")
    ```
  

- `str_extract()`: Returns the pattern that it finds. `str_extract()` will only
  return the first match but `str_extract_all()` will return all matches.
    ```{r}
    colorstr <- str_c("red", "blue", "yellow", "orange", "brown", sep = "|")
    colorstr
    str_extract("I like blue and brown and that's it", colorstr)
    str_extract_all("I like blue and brown and that's it", colorstr)
    ```
  
- `str_match()`: returns a matrix where each column is a grouped component.

    ```{r}
    x <- "I like blue and brown and that's it, or black"
    str_extract_all(x, "(and|or)\\s([^\\s]+)")
    str_match_all(x, "(and|or)\\s([^\\s]+)")
    ```
  
- Let's look at the poem "Farewell We Call to Hearth and Hall!"
    ```{r}
    farewell <- c("Farewell we call to hearth and hall!
                  Though wind may blow and rain may fall,
                  We must away ere break of day
                  Far over wood and mountain tall.")
    writeLines(farewell)
    ```

- `str_split()` will split up a string based on a character we choose.

    ```{r}
    ## Split based on spaces
    str_split(farewell, pattern = "\\s+", simplify = TRUE) ## use one or more space to split
    ```
    
- `str_replace()` and `str_replace_all()` will replace patterns with provided
  strings. So say we want to get rid of all punctuation.
  
    ```{r}
    str_split(farewell, pattern = "\\s+", simplify = TRUE) %>%
      str_replace_all("\\.|\\!|,", "")
    ```
    
- You can use back references to populate the replacement.

    ```{r}
    str_replace_all("It is 1am", "(\\d+)(am|pm)", "\\2")
    ```


- More stringr options can be found in [RDS](https://r4ds.had.co.nz/).


  

